---
title: "Lab 1: Hello R!"
date: 2025-10-07
---


```{r}
library(tidyverse)
library(palmerpenguins)
```

Look at the table

If you want to learn more about what is in the data, you can use the fucntion `str()` (for "structure"). As before, replace  `____` with the name of the data frame:


```{r}
str(____)
```

You will see a formal description of each column in the data frame. It tells you about the data type, which will be important later.

Generally, the data can either be numbers, or text. Numbers can either be "whole" numbers, or integers (abbrevated `int` in R), or numbers with a comma (abbrevated `num`) in R.

```{r}
penguins <- penguins
```


```{r}
penguins %>%
  group_by(species, sex) %>%
  summarise(mean_bill_length_mm = mean(bill_length_mm))
```


```{r}
penguins_summary <- penguins %>%
  filter(!is.na(sex)) %>%
  group_by(species, sex) %>%
  summarise(
    mean_bill_length_mm = mean(bill_length_mm),
    min_bill_length_mm = min(bill_length_mm),
    max_bill_length_mm = max(bill_length_mm),
    sd = sd(bill_length_mm)
  )

penguins_summary %>%
  ggplot(aes(x = species)) +
  geom_bar(
    aes(y = mean_bill_length_mm, fill = sex),
    stat = "identity",
    position = position_dodge()
  ) +
  geom_errorbar(
    aes(
      ymin = mean_bill_length_mm - sd,
      ymax = mean_bill_length_mm + sd,
      fill = sex
    ),
    position = position_dodge()
  ) +
  labs(x = "Species", y = "Body Mass (g)", fill = "Sex") +
  scale_fill_manual(values = c("darkblue", "lightblue")) +
  theme_classic()

```

calcultate things

::: {.column-screen-inset}
pebf
:::

```{r}
penguins_new <- penguins %>%
  mutate(bill_ratio = bill_depth_mm / bill_length_mm) %>%
  group_by(species, sex) %>%
  summarise(
    mean_bill_ratio = mean(bill_ratio),
    mean_body_mass_g = mean(body_mass_g)
  )
```


```{r}
penguins_new %>%
  filter(!is.na(sex)) %>%
  ggplot(aes(x = species, y = mean_body_mass_g, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Species", y = "Body Mass (g)", fill = "Sex") +
  scale_fill_manual(values = c("darkred", "lightblue")) +
  theme_classic()
```


```{r}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, col = species)) +
  geom_point() +
  theme_classic()
```


```{r}
mod1 <- glm(
  flipper_length_mm ~ body_mass_g * species * sex,
  data = penguins,
  family = Gamma(link = "log")
)

mod2 <- glm(
  flipper_length_mm ~ body_mass_g + species + sex,
  data = penguins,
  family = Gamma(link = "log")
)


AIC(mod1, mod2)


```


```{r}
library(broom)
library(marginaleffects)
predictions(mod1, condition = c("species", "sex"))

m <- glm(
  flipper_length_mm ~ species * sex,
  data = penguins,
  family = Gamma(link = "log")
)

avg_comparisons(blubb, variables = list(species * sex = "pairwise"))

avg_predictions(
  mod1,
  by = c("sex", "species"),
  newdata = "balanced",
  hypothesis = difference ~ sequential
)

avg_slopes(mod1, variables = c("sex"), by = "species")

```

```{r}

```

```{r}
augment(
  mod2,
  penguins %>%
    filter(!is.na(sex)),
  type.predict = "response",
  se_fit = T
) %>%
  ggplot() +
  geom_point(
    aes(
      y = flipper_length_mm,
      x = body_mass_g,
      col = sex,
      shape = sex
    ),
    alpha = .8
  ) +
  geom_ribbon(
    aes(
      ymin = .fitted - .se.fit,
      ymax = .fitted + .se.fit,
      x = body_mass_g,
      fill = sex,
      linetype = sex
    ),
    alpha = .3
  ) +
  geom_line(aes(y = .fitted, x = body_mass_g, col = sex, linetype = sex)) +
  facet_wrap(~species) +
  theme_classic()

```


```{r}
#| fig.width: 8
#| fig.height: 6
#| dpi: 150

library("performance")

#check_model(mod1)

r2(mod1)
r2(mod2)

compare_performance(mod1, mod2)
```


```{r}
library(marginaleffects)

plot_predictions(
  mod1,
  condition = c("body_mass_g", "species", "sex"),
  points = .4
) +
  ggtitle("Interactions") +
  facet_grid(sex ~ species, scales = "free") +
  theme_light()

plot_predictions(
  mod2,
  condition = c("body_mass_g", "species", "sex"),
  points = .4
) +
  ggtitle("Additive") +
  facet_grid(sex ~ species, scales = "free") +
  theme_light()
```