---
title: "Indications pour TP 01"
format: html
date: 2025-12-01
---

Here I give you some help in case you struggle solving the tasks. Please try to solve the tasks first without these hints, all information you need is in the `.qmd` document I shared with you.


:::: project
### Task 1

Use the R code chunk below and replace `____` with the name of the data frame (`penguins`). When you run the chunk, you will see the first rows:

```{r}
#| eval: false
#| echo: true

head(____)
```

::: {.callout-tip collapse="true" title="Show Hint"}
`head()` is a function. To use it, you put the name of the data set inside the parentheses. The name of the dataset is `penguins`, so you want to replace `____` with that name.
:::
::::

:::: project
### Learning Check 1

**What does each row in `penguins` contain?**

A.  Data for all penguins at a specific island, e.g. Biscoe

B.  Data for all penguins of a species, e.g. Adelie

C.  Data for only one penguin

::: {.callout-tip collapse="true" title="Show Hint"}
Try to imagine how the researchers entered the data. What would be the natural unit of observation for field biologists?
:::
::::

:::: project
### Task 2

Select the column `bill_length_mm` from the `penguins` data:

```{r}
#| eval: false
#| echo: true

____$____
```

::: {.callout-tip collapse="true" title="Show Hint"}
The name of the dataset is `penguins`, and you want to access a column called `bill_length_mm`. So first, you tell R the name of the data, then, that you want to access a column (`$`), and then you give the name of the column (`bill_length_mm`).
:::
::::

:::: project
### Task 3

Calculate the minimum, maximum, and mean value for `bill_length_mm`:

```{r}
#| eval: false
#| echo: true

min(____$____)
max(____$____)
mean(____$____)
```

::: {.callout-tip collapse="true" title="Show Hint"}
Above, you learned how to access a column in a data frame with the `$` operator. Use the same method here, i.e. Insert in each function first the name of the data frame (`penguins`), then use `$` to access the `bill_length_mm` column.
:::
::::

:::: project
### Task 4

Group the data by `species` and `sex` before calculating the mean of `bill_length_mm`:

```{r}
#| eval: false
#| echo: true

penguins %>% # 1. Take the penguins dataset
  group_by(species, ____) %>% # 2. Group it by species and sex
  summarize(mean_bill_length_mm = mean(bill_length_mm)) # 3. create a new column called "mean_bill_length_mm" and assign the mean values of each group
```

::: {.callout-tip collapse="true" title="Show Hint"}
`group_by()` works with multiple grouping variables. You can just add `sex` to group the data additionally by sex. The mean (step 3) will then be calculated for the combinations of both groups.
:::
::::

:::: project
### Task 5

Group the data by `species` and `sex` and additionally to the mean and minimum value of `bill_length_mm`, calculate the maximum value. **Hint:** If you create columns in R, they cannot contain spaces or minuses `-`.

```{r}
#| eval: false
#| echo: true

penguins %>% # 1. Take the penguins dataset
  group_by(species, ____) %>% # 2. Group it by species and sex
  summarize(
    mean_bill_length_mm = mean(bill_length_mm), # 3.1. create a new column called "mean_bill_length_mm" and assign the mean values of each group
    min_bill_length_mm = min(bill_length_mm), # 3.2. create a new column called "min_bill_length_mm" and assign the minimum values of each group
    ____ = ____(____) # 3.3. create a new column called "max_bill_length_mm" and assign the maximum values of each group
  )
```

::: {.callout-tip collapse="true" title="Show Hint"}
As in Task 4, group the data additionally by `sex`. In step 3.3, you can now calculate the maximum value of `bill_length_mm`. The name on the left of `=` is the name of the new column you will create. Call this column `max_bill_length_mm`. Remember, you create a new column before the `=`. Then use the `max()` function to calculate the maximum value of `bill_length_mm`.
:::
::::

:::: project
### Learning Check 2

Why do you think it is dangerous to do the following:

```{r}
#| eval: false
#| echo: true

penguins <- penguins %>%
  filter(species == "Adelie")
```

::: {.callout-tip collapse="true" title="Show Hint"}
With the filter function you only retain a subset of the data. You would overwrite the entire dataset with only one species. Think about what would happen in later tasks.
:::
::::

:::: project
### Task 6

Create a new subset of the data. Use a reasonable name for the new data set. Filter to exclude Adelie penguins (with `!=`, see above), and a `body_mass_g` smaller or equal to 4000 g:

```{r}
#| eval: false
#| echo: true

____ <- ____ %>%
  filter(____, ____)
```

::: {.callout-tip collapse="true" title="Show Hint"}
First, select a reasonable name for the new subset of data you are creating. This name should reflect how you filtered the original data. `penguins_chinstrap_entoo_light`, `peng_not_adelie_below4kg`, etc. would work.

Next (after the `<-`), you have to give the data set you want to filter (`penguins`).

In the filter function, you want to first exclude all Adelie penguins. Think about this as "entries in column `species` is not equal to (`!=`) Adelie". Remember that text in R needs to be wrapped with `""`.

As the second filtering step, only keep penguins below or equal to 4000 g. Think about this as "entries in column `body_mass_g` is smaller or equal to (`<=`) 4000 g". Remember that numbers don't need to be wrapped with `""`.

Remember that each condition inside `filter()` is separated by a comma.

:::
::::

:::: project
### Task 7

Use `str()` for the `penguins` data set:

```{r}
#| eval: false
#| echo: true

str(____)
```

::: {.callout-tip collapse="true" title="Show Hint"}
`str()` shows the structure of a data frame: variable types, lengths, etc. You have to insert the name of the data set (`penguins`) into the `str()` function.
:::
::::

::: project
### Task 8

Use `levels()` below, to see the levels of `penguins$island`:

```{r}
#| eval: false
#| echo: true

levels(____$____)
```

::: {.callout-tip collapse="true" title="Show Hint"}
`levels()` shows all possible categories in a factor. This is useful before plotting or modelling categorical variables. You have to insert the name of the data set and of the column into the `levels()` function. Remember that you can access a column with the `$` operator.
:::
:::

