---
title: "TP 06 : Distributions statistiques alternatives"
format: html
editor: visual
---

```{r}
library(here)
library(tidyverse)
library(performance)
library(modelbased)
library(see)
library(marginaleffects)
```

## Coral Recruitment

This data comes from CRIOBE surveys measuring coral recruitment (i.e., larvae that settle). Recruitment was measured by placing terracotta tiles (approximately 15 × 15 cm) at three different sites as settlement substrates. Twenty tiles were deployed per site. After one year, the number of coral recruits that settled and survived on each tile was counted. The data range from 0 to 13 recruits per tile.

Data structure:

- **Response variable**: Number of coral recruits per tile (count data)
- **Explanatory variable**: Site (3 levels)
- **Sample size**: 60 tiles total (20 per site)

```{r}
dat_recruits <- read.csv(here("TP/TP_06/data/recruitment.csv"))
```

Let's have a look at the data

```{r}
ggplot(data = dat_recruits, aes(x = site, y = recruits, col = site)) +
  geom_jitter(size = 3, alpha = .6, height = 0) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Number of recruits per tile") +
  scale_color_ordinal()
```

### Task 1

Do you think that count data like this can be analysed using a normal distribution?

Why not? Consider the following:

- What is the minimum possible value? Can we observe negative recruit counts?
- How is the variance distributed relative to the mean?
- Is the relationship between mean and variance linear or does it change?

What would be the appropriate distribution for count data? (You can refer to the [GLM overview](https://andieich.github.io/2026_biostatistiques_2/distr_overview.html) for guidance on distribution families.)

Now let's compare two models: one assuming a normal distribution (which is incorrect for count data) and one using the appropriate distribution.

**Model 1: Normal distribution (incorrect)**

```{r}
lm_recruits <- lm(recruits ~ site, data = dat_recruits)

check_model(lm_recruits)
```

Look at the diagnostic plots, especially:

- **"Normality of residuals"**: Does the distribution look normal? What problems do you notice?
- **"Homogeneity of variance"**: Are the residuals equally spread across fitted values?
- **First plot (residuals vs fitted)**: Can the model reliably predict negative recruit counts?

### Task 2

**Model 2: Using the appropriate distribution**

Now fit the correct model by replacing the `____` in `family =` with the appropriate distribution. Choose from:

- `poisson()` - for count data with equal mean and variance
- `Gamma()` - for continuous positive data
- `binomial()` - for binary (yes/no) data

```{r}
glm_recruits <- glm(recruits ~ site, data = dat_recruits, family = ____)

check_model(glm_recruits)
```

### Task 3

You can use the `compare_performance()` function (`modelbased` package) to compare the performance of both models:

```{r}
compare_performance(lm_recruits, glm_recruits)
```

Have a look at the R² values (do you remember what they tell you?). For for the "normal" linear model, look at column "R2", for the GLM, look at column "Nagelkerke's R2".

Another useful metric is the AIC (Akaike Information Criterion), which also describes the quality of a model. The lower the value, the better!

Which of the two model fits better and is more suitable for the type of data?

### Task 4

**Analyzing the results**

```{r}
anova(glm_recruits)
```

Is there a significant difference in recruitment rates among the three sites? What is the p-value, and what does it mean biologically (e.g. which site is more suitable for recruitment? It could e.g. be prioritised for MPAs)?

### Task 5

**Visualizing predictions**

Let's plot the model-predicted means and 95% confidence intervals with the raw data overlaid.

**Hint:** I just found a new functionality in the `modelbased` plot function we were using to overlay the raw data (with `show_data = TRUE`). I also turn off the lines that connected the estimated means (with `join_dots = FALSE`), since to me, this implies a numerical explanatory variable, but this is personal choice.

```{r}
estimate_means(glm_recruits) %>%
  plot(show_data = TRUE, join_dots = FALSE) +
  theme_light() +
  labs(title = "Predicted coral recruitment by site")
```

**Post-hoc comparisons**

Now let's formally test which sites differ significantly from each other:

```{r}
estimate_contrasts(glm_recruits)
```

Based on the post-hoc tests:

- Which pairs of sites differ significantly?
- Which pairs do not differ significantly?
- If you could only protect two of the three sites, which would you chose?

**For publication**: You would typically report the ANOVA table (with p-value), the plot with predictions and raw data, the R² value, and place the post-hoc contrast table in an appendix.

## Fish Biomass

This data comes from MCR LTER Annual Fish Surveys (Gump station on Moorea). We're examining Secondary Consumer biomass across three habitat types (Backreef, Forereef, Fringing) in 2024. Each observation represents the mean biomass from a single site-habitat combination.

Data structure:

- **Response variable**: Mean Secondary Consumer biomass per site (g)
- **Explanatory variable**: Habitat type (3 levels)
- **Sample size**: 18 observations (6 sites × 3 habitats)

```{r}
fish_biomass <- read.csv(here("TP/TP_06/data/fish_biomass.csv"))
```

Let's visualise the data:

```{r}
ggplot(data = fish_biomass, aes(x = habitat, y = biomass, col = habitat)) +
  geom_jitter(size = 3, alpha = .6, width = 0.2) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Habitat", y = "Biomass (g)") +
  scale_color_ordinal()
```

### Task 6

Look at the plot and consider: - Can biomass values be negative?

- Does the variance (spread of points) appear constant across habitats, or does it increase with larger mean values?
- What does this tell you about whether a normal distribution would be appropriate?

**Model 1: Normal distribution (incorrect)**

Let's first try fitting a linear model with a normal distribution and see what problems arise:

```{r}
lm_fish <- lm(biomass ~ habitat, data = fish_biomass)

check_model(lm_fish)
```

Look at the diagnostic plots and consider:

- Can you see any predictions that might be negative (which is biologically impossible for biomass)?
- How do the residuals look? Are they normally distributed?
- Are the assumptions violated?

### Task 7

**Model 2: Using the appropriate distribution**

Now fit the correct model by replacing the `____` in `family =` with the appropriate distribution. Choose from:

- `poisson()` - for count data with equal mean and variance
- `Gamma()` - for continuous positive data
- `binomial()` - for binary (yes/no) data

```{r}
m_fish <- glm(
  biomass ~ habitat,
  data = fish_biomass,
  family = ____
)

check_model(m_fish)
```

Examine the diagnostic plots. How well do the assumptions appear to be met?

### Task 8

Compare the two models:

```{r}
compare_performance(lm_fish, m_fish)
```

Which model is more appropriate for this data and why?

### Task 9

**Analyzing the results**

```{r}
anova(m_fish)
```

Is there a significant difference in Secondary Consumer biomass among habitats? What is the p-value?

### Task 10

**Visualizing predictions**

Let's plot the model-predicted means and 95% confidence intervals with the raw data overlaid:

```{r}
estimate_means(m_fish) %>%
  plot(show_data = TRUE, join_dots = FALSE) +
  theme_light()
```

**Post-hoc comparisons**

Now let's formally test which habitats differ significantly from each other:

```{r}
estimate_contrasts(m_fish)
```

Which habitat has the highest predicted biomass? Which has the lowest?

## Coral Cover

The coral cover data we were using so far comes from point intercept transects. At each site, a transect tape was laid on the seafloor. For 150 point it was checked if there is coral below ("yes") or not ("no"), this was done for the three main genera *Acropora*, *Montipora*, and *Pocillopora*.

Read data:

```{r}
dat_coral <- read.csv(here("TP/TP_06/data/coral_2020.csv"))
```

Data structure:

- **Response variable**: Coral cover as count of points per genus (out of 150 points per site)
- **Explanatory variables**: Coast (3 levels) and Coral genus (3 levels), including interaction
- **Sample size**: 9 observations (3 coasts × 3 genera)

An plot raw data

```{r}
dat_coral %>%
  ggplot(aes(x = coast, y = points, col = coral_genus)) +
  geom_point(
    size = 3,
    position = position_jitterdodge(jitter.width = .2, dodge.width = .5)
  ) +
  theme_minimal() +
  labs(
    x = "Coast",
    y = "Number of points (out of 150)",
    color = "Coral genus"
  ) +
  scale_color_ordinal()
```

So far, we used the percent data calculated knowing that there are 150 points per site. We can calculate the cover in percent like this:

```{r}
dat_coral <- dat_coral %>%
  mutate(percent = 100 * (points / 150))
```

### Task 11

Why are linear models not suitable for this type of percent data? Think about the possible values your response variable could have and any limiting values.

To see the implications, we can still fit a linear model.

**Model 1: Normal distribution (incorrect)**

```{r}
lm_coral <- lm(percent ~ coast * coral_genus, data = dat_coral)

check_model(lm_coral)
```

Look at the diagnostic plots. The diagnostics might look relatively reasonable, but consider: **What are the predictions from this model? Can percentages be negative or exceed 100%?**

Let's visualise the predictions:

```{r}
estimate_means(lm_coral) %>%
  plot(show_data = TRUE, join_dots = FALSE) +
  theme_light()
```

Look carefully at the plot. Can you see predictions that are biologically impossible? Where would negative coral cover appear on this plot?

### Task 12

**Model 2: GLM Using the appropriate distribution**

Let's now fit a model with the appropriate distribution. The most honest way to do so, so is by using the raw data, i.e. here, the points (coral: yes or no?) per transect. The precision to measure the cover differs between transects with only 10 points, or transects with 150 points. In binomial models, we can include this by using `cbind(success, failure)`. In the coral example, `success` is the number of points per transect with coral (column name `points`) and failure the number of points per transect without coral (150 - `points`).

Now fit the correct model by replacing the `____` in `family =` with the appropriate distribution. Choose from:

- `poisson()` - for count data with equal mean and variance
- `Gamma()` - for continuous positive data
- `binomial()` - for binary (yes/no) data

If you decide for a binomial model, use `cbind(points, 150 - points)` as response, otherwise, the number of points (but think about if there is an upper limit of points per transect), or the calculated percent values.

```{r}
glm_coral <- glm(
  ____ ~ coast * coral_genus,
  data = dat_coral,
  family = ____
)

check_model(glm_coral)
```

I hope you chose a binomial model here. In that case, `compare_performance()` does not work, since the response variable differs between both model (LM: percent, GLM: success vs. failure). But the decision for the correct model can here be done by thinking about the nature of the data.

We can also plot the predicted values of both model to see if a model can predict values that are not possible (here, percent values below 0% or above 100%).

Let's do so for the LM

```{r}
estimate_means(lm_coral) %>%
  plot(join_dots = FALSE) +
  theme_light()
```

And for the GLM

```{r}
estimate_means(glm_coral) %>%
  plot(join_dots = FALSE) +
  theme_light()
```

### Task 13

What differences do you notice between both plots models:

- Are the confidence intervals different?
- Do all predictions fall between 0 and 100%?

### Task 14

**Analyzing the results**

```{r}
anova(glm_coral)
```

Is there a significant interaction between coast and genus? What do the p-values tell you?

### Task 15

**Post-hoc comparisons**

Let's examine which genera differ significantly within each coast:

```{r}
estimate_contrasts(glm_coral, p_adjust = "tukey", by = "coast")
```

Based on the contrasts and the visualisation for the GLM you did further above

- Which genera differ significantly within each coast?
- Do the significant differences change between coasts?
- How do these results compare to what the LM would suggest?

### Task 16

For comparison, here are the LM contrasts:

```{r}
estimate_contrasts(lm_coral, p_adjust = "tukey", by = "coast", )
```

Notice the differences in which groups are considered as significantly different. This highlights why using the correct distribution matters - the statistical inference can change!