---
title: "TP 04-2: Model Assumptions"
format: html
editor: visual
---

```{r}
#| echo: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(here)
library(performance)
```

# 2nd Part: Assumptions

This was all fun, but so far we don't know if the models are "valid" and if the three assumptions are met.

For the three models we created in the first part of the TP:


-   `m_poci`

-   `m_cover`

-   `m_ancova`


We can check if the assumptins are met ot notö

Use the `performance` package and the `check_model()` function and visually assess the models we used so far


```{r}
m_poci <- lm(percent_cover ~ coast, data = dat_coral_cover_poci)
check_model(m_poci)
```

```{r}
m_cover <- lm(percent_cover ~ genus * coast, data = dat_coral_cover_genus)
check_model(m_cover)

```

```{r}
m_ancova <- lm(flipper_length_mm ~ body_mass_g * species, data = penguins)
check_model(m_ancova)

c <- plot(check_model(m_ancova))
ggsave(
  c,
  filename = "check_model3.png",
  path = here("TP/TP_04/images"),
  width = 20,
  height = 20,
  units = "cm"
)
```

------------------------------------------------------------------------

## Part 3: Model Assessment with `performance` Package

Once you've built a model, you should check whether it meets the assumptions of linear regression. The `performance` package makes this easy!

### Task 10: Run full model diagnostics

Use `check_model()` to assess all assumptions at once:

```{r}
check_model(____)
```

This produces several diagnostic plots:

1.  **Linearity** (top-left): Points should be randomly scattered around the line. If they follow a pattern (like a smile), the relationship might not be linear.

2.  **Homogeneity of variance** (top-right): Residuals should have constant spread. If the band gets wider or narrower, variance isn't constant.

3.  **Normality** (middle): Points should follow the diagonal line. Deviations at the tails suggest non-normal residuals.

4.  **Posterior predictive check** (lower panels): Shows if the model can reproduce the data well.

### Task 11: Focus on homogeneity specifically

To zoom in on whether residual variance is constant, use:

```{r}
plot(check_model(____, check = "homogeneity"))
```

**Interpretation:** - Look at the **"Standardized Residuals vs Fitted"** plot - If you see a **cone shape** (wider on the right or left), that's heteroscedasticity - If the scatter is **roughly uniform** width, that's good (homoscedasticity)

### Task 12: Quick assessment questions

Examine the diagnostic plots and answer:

1.  Do the residuals appear normally distributed?
2.  Is the variance roughly constant across fitted values?
3.  Are there any obvious outliers or concerning patterns?
4.  Overall, do you think the model meets the linear regression assumptions?

------------------------------------------------------------------------

## Bonus Task

### Challenge - Build your own analysis

Choose **another continuous response variable** (e.g., `flipper_length_mm`, `bill_length_mm`, `bill_depth_mm`) and repeat the ANCOVA analysis:

1.  Create a scatter plot with fitted lines for each species
2.  Build the interaction model
3.  Check if the interaction is significant
4.  Compare slopes using `estimate_contrasts()`
5.  Run model diagnostics with `check_model()`

Write your code below:

```{r}
# Step 1: Visualization

# Step 2: Model

# Step 3: ANOVA

# Step 4: Post-hoc contrasts

# Step 5: Diagnostics

```

**Questions to answer:** - Is there an interaction between your chosen variable and species? - Do the species differ significantly in their response? - Do the diagnostics suggest any violations of model assumptions?