---
title: "Shining a Light on Daytime Coral Spawning Synchrony Across Oceans"
subtitle: "A GLM Case Study"
format: html
---


## Context

This analysis explores coral (*Porites rus*) spawning timing patterns across different habitats and islands in French Polynesia. We'll investigate whether spawning behavior varies by environmental factors using generalized linear models.

**Reference**: Moritz et al. (2025). Shining a light on daytime coral spawning synchrony across oceans. *Global Ecology and Biogeography*, 10.1111/geb.70072

**Data source**: SEANOE (10.17882/105886)


## Load Packages and Data

Load packages needed

```{r}
library(tidyverse)
library(here)
library(performance)
library(modelbased)
```

Load `dat_sites`. This data set gives an overview of spawning at different parts of the word:

```{r}
dat_sites <- read_csv(here(
  "lectures/07/Porites_rus_spawning/data/clean/prus_sites.csv"
))
```

## Figure 1B


Have a look at the data set: Which columns exist? Remember, you can use `str()` to get an overview or click on the table in the environment tab to the right.

No, let's recreate Figure 1B of the publication. We only focus on the exact data.

1.  Make a new data set and filter it for the "Lagoon" habitat (`filter()`)
2.  Create a summary data frame by grouping your lagoon data set by the column `island` and calculate with `summary()` the mean and standard deviation (`sd()`) of `h_after_sunrise` (the hours after sunrise).

```{r}
dat_lagoon <- dat_sites %>%
  filter(habitat == "Lagoon")

dat_lagoonS <- dat_lagoon %>%
  group_by(island) %>%
  summarise(
    mean_h_after_sunrise = mean(h_after_sunrise),
    sd_h_after_sunrise = sd(h_after_sunrise)
  )
```

3.  Now the data is ready to be plotted. Plot the summarised data using `ggplot()` with the following configuration:

    - `island` on x axis
    - `geom_pointrange()` layer with the mean as `y`, and mean - SD as `ymin` and mean + SD as `ymax`
    - add a horizontal line (`geom_hline()`) layer with the mean over all the mean values in your summary data frame as `yintercept`.

```{r}
dat_lagoonS %>%
  ggplot(aes(x = island)) +
  geom_pointrange(
    aes(
      y = mean_h_after_sunrise,
      ymin = mean_h_after_sunrise - sd_h_after_sunrise,
      ymax = mean_h_after_sunrise + sd_h_after_sunrise
    ),
    col = "royalblue3"
  ) +
  labs(y = "Hours after sunrise", x = NULL) +
  geom_hline(yintercept = mean(dat_lagoonS$mean_h_after_sunrise)) +
  theme_test(base_size = 16) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

What do you see?

## Figure 2A

Now, lets try to better understand the variability by recreating Fig. 2A. Spawning is often synchronised by light intensity. Therefore above, the hours after sunrise were used. Spawning time may also be impacted by the habitat, i.e. Channel, Lagoon, or Outer slope. Now, let's verify this. Plot the start time (`h_start`, in "decimal hours") on the y axis against the time of sunrise on the y axis (`h_sunrise`).

The data was collected at different habitats: Channel, Lagoon, Outer Slope. To see how this impacted spawning time use different colours for the different habitats in column `habitat`.

```{r}

col_hab <- c(
  "Channel" = "#401801",
  "Lagoon" = "#3A848C",
  "Outer slope" = "#465E8C"
)

dat_sites %>%
  ggplot(aes(y = h_start, x = h_sunrise, fill = habitat)) +
  geom_point(shape = 21) +
  labs(y = "Time start", x = "Time sunrise", fill = "Habitat") +
  scale_fill_manual(values = col_hab) +
  theme_test(base_size = 16)
```

This calls for a regression! 

With an ANCOVA, we can see if the relationship differs between "Lagoon" and "Outer slope". There is only one observation for "Channel", so we have to ignore it. First, create another data set without (`!=`) the Channel habitat using `filter()`:

```{r}
dat_sites_sel <- dat_sites %>%
  filter(habitat != "Channel")
```

Now, let's do a linear model, using `h_start` as response variable and the interaction of `h_sunrise` and `habitat` as explanatory variable. 

```{r}
lm_sr_hab_int <- lm(h_start ~ h_sunrise * habitat, data = dat_sites_sel)
```

Do the actual analysis. Use `anova()` to see if `h_sunrise`, `habitat`, and their interaction have a significant impact:

```{r}
anova(lm_sr_hab_int)
```

Which explanatory varaibles have a signficant impact on `h_start`?

What do you think, is it useful to keep the interaction in the model?

Now, let's check the assumptions with `check_model()`:

```{r}
check_model(lm_sr_hab_int)
```

Do you think that the residuals are normally distributed?

Now let's make a second model without the interaction and see if the residuals are normally distributed:

```{r}
lm_sr_hab <- lm(h_start ~ h_sunrise + habitat, data = dat_sites_sel)
check_model(lm_sr_hab)
```

That still does not look good!

But wait, all observations are positive, time cannot be negative! Which type of GLM would you use?

Use `check_model()` to see if the residuals fit better to the distribution you chose. 

First, use the interaction as explanatory variable, it could be that the change of the distributions leads to a change in p-values. If the interacion is not significant, I would use a model with the additive efects of `h_sunrise` and `habitat`. 

```{r}
glm_sr_hab_int <- glm(
  h_start ~ h_sunrise * habitat,
  data = dat_sites_sel,
  family = Gamma()
)

anova(glm_sr_hab_int)
```


```{r}
glm_sr_hab <- glm(
  h_start ~ h_sunrise + habitat,
  data = dat_sites_sel,
  family = Gamma()
)

anova(glm_sr_hab)

```


Which model would you use in a publication? The linear model or the GLM? Interaction or addititve effects?

We can now add the regression lines and the confidence intervalls to Fig 2A. To do so, first create a data frame with the model predictions using `estimate_relation()` and the model you chose:

```{r}
pred_glm_sr_hab <- estimate_relation(glm_sr_hab)
```

We can now add this to the figure we created further above. Have a look at the prediction dataframe you just created. What column should be used for the regression line (in a `geom_line()` layer), what is is columns defining the confidence intervals (in a `geom_ribbon()` layer)? Add them to the ggplot function. Remember to use the summary data for these two layers.

```{r}
dat_sites %>%
  ggplot(aes(y = h_start, x = h_sunrise, fill = habitat)) +
  geom_ribbon(
    data = pred_glm_sr_hab,
    aes(ymin = CI_low, ymax = CI_high, y = NULL),
    alpha = .3,
    show.legend = F
  ) +
  geom_line(
    data = pred_glm_sr_hab,
    aes(y = Predicted, col = habitat),
    show.legend = F
  ) +
  geom_point(shape = 21) +
  labs(
    y = "Time start",
    x = "Time sunrise",
    fill = "Habitat",
    col = "Habitat",
  ) +
  scale_colour_manual(values = col_hab) +
  scale_fill_manual(values = col_hab) +
  theme_test(base_size = 16)
```

So, as a summary, so far we learned that the time of spawing is around 1.6 h after sunrise. The spawning time is significantly correlated to the time of sunrise. This means that the later the sunrise, the later the spawnign time. This relation does not differ betwen Outer slope and Lagoon, although the spawning time is much later in the Outer slope than in the Lagoon. We can formally test this with a post-hoc test using `estimate_contrasts()`:

```{r}
estimate_contrasts(glm_sr_hab, p_adjust = "tukey")
```

Look what the p-value actually means, try to understand the yellow text: 
"Predictors averaged: h_sunrise (5.5)". At which point on the regression line were the two habitats compared? How much later is the spawning on average in the Outer reef?

To simplify things, we can concentrate on the the time after sunrise (`h_after_sunrise`), since we know that the spawnign time and sunrise is significantly correlated. This lets us visualise the difference like this:
  
```{r}
dat_sites %>%
  ggplot(aes(y = h_after_sunrise, x = habitat, fill = habitat)) +
  geom_violin(alpha = .7, col = NULL) +
  geom_jitter(shape = 21, width = .2, alpha = .7) +
  labs(y = "Hours after sunrise", x = NULL) +
  coord_flip() +
  scale_fill_manual(values = col_hab) +
  theme_test(base_size = 16) +
  theme(legend.position = "none")
```


So let's test differences here with ANOVA, ie comparison of groups:

Use the same dataset from the regression above (i.e. excluding the single Channel observation) and make a model using `h_after_sunrise` as response and `habitat` as explanatory variable. Consider what you learned above about a suitable distribution. Do this yourself:

1.  Make model
2.  Check assumptions with `check_model()`
3.  Test if habitats differ significantly with `anova()`
4.  Predict data using `estimate_means()`
5.  Add the estimated mean and CIs as a `geom_pointrange()` layer


```{r}
glm_hab <- glm(
  h_after_sunrise ~ habitat,
  data = dat_sites_sel,
  family = Gamma()
)

check_model(glm_hab)

anova(glm_hab)

pred_glm_hab <- estimate_means(glm_hab)

dat_sites %>%
  ggplot(aes(y = h_after_sunrise, x = habitat, fill = habitat)) +
  geom_violin(alpha = .7, col = NULL) +
  geom_jitter(shape = 21, width = .2, alpha = .7) +
  geom_pointrange(
    data = pred_glm_hab,
    shape = 21,
    col = "grey30",
    aes(y = Mean, ymin = CI_low, ymax = CI_high, col = habitat)
  ) +
  labs(y = "Hours after sunrise", x = NULL) +
  coord_flip() +
  scale_fill_manual(values = col_hab) +
  scale_colour_manual(values = col_hab) +
  theme_test(base_size = 16) +
  theme(legend.position = "none")
```


As before, you can use `estimate_contrast()` to calculate the difference of the spawning time between Lagoon and Outer slope:

```{r}
estimate_contrasts(glm_hab)
```

Why does is slightly differ to the difference calcualted above?


## Figure 2B

The spawning time may also be impacted by the depth. Let's plot it as in Fig. 2B. Use `depth` as x axis, `h_after_sunrise` as y axis, and use different colours for the habitats.

```{r}
dat_sites %>%
  ggplot(aes(y = h_after_sunrise, x = depth, fill = habitat)) +
  geom_jitter(shape = 21, width = .2, alpha = .7) +
  scale_fill_manual(values = col_hab) +
  labs(y = "Hours after sunrise", x = "Depth (m)", fill = "Habitat") +
  theme_test(base_size = 16)
```


Would you do a regression as in the publication? Why not? Think about indendence of the data! Do you think the habitat could have an impact besides the depth? Is it possible to consider the impact of habitat and depth?

## Figure 4B

To better understand the variablity, lets investigate the impact of sex. The authour provided a more detailed dataset for this analysis for coral spawning observaitions in Vairao. Load it and have a look at it:

```{r}
dat_colony <- read_csv(here(
  "lectures/07/Porites_rus_spawning/data/clean/prus_colony.csv"
))
```


Now, plot it as follows:

```{r}
ggplot(data = dat_colony, aes(x = sex, y = h_after_sunrise, fill = sex)) +
  geom_violin(alpha = .7, col = NULL) +
  geom_jitter(shape = 21, width = .2, alpha = .7) +
  labs(y = "Hours after sunrise", x = NULL) +
  scale_fill_manual(values = c("#FFA500", "#551A8B")) +
  coord_flip() +
  theme_test(base_size = 16) +
  theme(legend.position = "none")
```

How would you analyse this data? What type of model woudl you use? Try it out, you know the process by now:

```{r}
glm_sex <- glm(h_after_sunrise ~ sex, data = dat_colony, family = Gamma())

check_model(glm_sex)

anova(glm_sex)

pred_glm_sex <- estimate_means(glm_sex)
```

If you're happy with the model, predict the results and add mean and CI to the visualistaion from above:

```{r}

col_sex <- c("MALE" = "#551A8B", "FEMALE" = "#FFA500")

ggplot(data = dat_colony, aes(x = sex, y = h_after_sunrise, fill = sex)) +
  geom_violin(alpha = .7, col = NULL) +
  geom_jitter(shape = 21, width = .2, alpha = .7) +
  geom_pointrange(
    data = pred_glm_sex,
    shape = 21,
    col = "white",
    aes(y = Mean, ymin = CI_low, ymax = CI_high),
  ) +
  labs(y = "Hours after sunrise", x = NULL) +
  scale_fill_manual(values = col_sex) +
  coord_flip() +
  theme_test(base_size = 16) +
  theme(legend.position = "none")
```

How much time lays between the mean male and femal spawning start? Use `estimate_contrasts()`:

```{r}
estimate_contrasts(glm_sex, p_adjust = "tukey")
```

## Figure 2F

As you see, there is a difference, but still a lot of variablitiy. Maybe that has to do with the water temperature (`sst`)? As always, start with a plot. Use `sst` as x axis , `h_after_sunrise` as y axis, and different colours for `sex`:

```{r}
ggplot(data = dat_colony, aes(x = sst, y = h_after_sunrise, fill = sex)) +
  geom_point(shape = 21) +
  labs(y = "Hours after sunrise", x = "SST (°C)", fill = "Sex") +
  scale_fill_manual(values = col_sex) +
  theme_test(base_size = 16)
```  

  
How would you test what you see?


```{r}
glm_sst_sex_int <- glm(
  h_after_sunrise ~ sst * sex,
  data = dat_colony,
  family = Gamma()
)

check_model(glm_sst_sex_int)

anova(glm_sst_sex_int)


glm_sst_sex <- glm(
  h_after_sunrise ~ sst + sex,
  data = dat_colony,
  family = Gamma()
)

check_model(glm_sst_sex)

anova(glm_sst_sex)
```


If you're happy with your model, add the regressio line and CI to the plot above. Do so as before by first creating a data frame with the predictions using `estimate_relation()`:

```{r}
pred_glm_sst_sex <- estimate_relation(glm_sst_sex)

ggplot(data = dat_colony, aes(x = sst, y = h_after_sunrise, fill = sex)) +
  geom_ribbon(
    data = pred_glm_sst_sex,
    aes(y = NULL, ymin = CI_low, ymax = CI_high),
    show.legend = F,
    alpha = .3
  ) +
  geom_point(shape = 21) +
  geom_line(
    data = pred_glm_sst_sex,
    aes(y = Predicted, col = sex),
    show.legend = F
  ) +
  labs(y = "Hours after sunrise", x = "SST (°C)", fill = "Sex") +
  scale_fill_manual(values = col_sex) +
  scale_colour_manual(values = col_sex) +
  theme_test(base_size = 16)
```


And how much does the spawnign time differ between male and female corals?

```{r}
estimate_contrasts(glm_sst_sex)
```

```{r}
estimate_contrasts(
  glm_sst_sex,
  by = "sst = c(27, 28, 29)"
)
```


## References

Moritz C, et al. (2025). Shining a light on daytime coral spawning synchrony across oceans. *Global Ecology and Biogeography*, **34**(3):e70072. https://doi.org/10.1111/geb.70072

Data: DOI 10.17882/105886