---
title: "Shining a Light on Daytime Coral Spawning Synchrony Across Oceans"
subtitle: "A GLM Case Study"
format: html
---

## Context

This analysis explores coral (*Porites rus*) spawning timing patterns across different habitats and islands in French Polynesia. We'll investigate whether spawning behavior varies by environmental factors using generalized linear models.

**Reference**: Moritz et al. (2025). Shining a light on daytime coral spawning synchrony across oceans. *Global Ecology and Biogeography*, 10.1111/geb.70072

**Data source**: SEANOE (10.17882/105886)

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(performance)
library(modelbased)
library(see)
library(lme4)
library(here)
library(lmerTest)

# Set theme
theme_set(theme_light(base_size = 12))
```

## Load and Prepare Data

```{r}
#| label: load-data

# Load site-level spawning data
spawn_data <- read.csv(
  here("lectures/07/Porites_rus_spawning/data/SEANOE_SiteSpawning_Moritz.csv"),
  sep = ";",
  dec = "."
) |>
  as_tibble() |>
  mutate(
    month = factor(
      month,
      levels = c("10", "11", "12", "1", "2", "3", "4"),
      labels = c(
        "October",
        "November",
        "December",
        "January",
        "February",
        "March",
        "April"
      )
    ),
    spawning = !no.start # Convert to TRUE/FALSE for spawning presence
  ) |>
  filter(site != "AME Teva i uta") # Remove unclear data

# Quick overview
glimpse(spawn_data)
```

## Exploratory Analysis

### Spawning Across Islands

First, let's see how many sites per island observed spawning events:

```{r}
#| label: spawning-by-island

spawn_data |>
  group_by(island) |>
  summarise(
    n_sites = n_distinct(site),
    n_spawning = sum(spawning, na.rm = TRUE),
    pct_spawning = round(100 * n_spawning / n(), 1)
  )
```

### Visualize Spawning Percentage by Island

```{r}
#| label: fig-spawn-percent
#| fig-cap: "Percentage of sites where spawning was observed by island"

spawn_summary <- spawn_data |>
  group_by(island) |>
  summarise(
    n_sites = n(),
    spawning_yes = sum(spawning, na.rm = TRUE),
    spawning_no = n_sites - spawning_yes,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = starts_with("spawning"),
    names_to = "status",
    values_to = "count"
  ) |>
  mutate(status = ifelse(status == "spawning_yes", "yes", "no"))

ggplot(spawn_summary, aes(x = island, y = count, fill = status)) +
  geom_col() +
  scale_fill_manual(values = c("grey20", "grey70"), labels = c("no", "yes")) +
  labs(y = "Number of sites", x = NULL, fill = "Spawning observed") +
  theme(legend.position = "bottom")
```

### Spawning Time by Habitat

Focus on observations where spawning was recorded (exact or approximate time):

```{r}
#| label: data-with-spawning

spawn_obs <- spawn_data |>
  filter(spawning == TRUE) |>
  mutate(
    habitat = factor(
      habitat,
      levels = c("Lagoon", "Channel", "Inner slope", "Outer slope")
    )
  )

# Summary statistics by habitat
spawn_obs |>
  group_by(habitat) |>
  summarise(
    n = n(),
    mean_time = mean(start.decimal, na.rm = TRUE),
    sd_time = sd(start.decimal, na.rm = TRUE),
    mean_hours_after_sunrise = mean(STrSR, na.rm = TRUE),
    sd_hours_after_sunrise = sd(STrSR, na.rm = TRUE),
    .groups = "drop"
  )
```

### Visualize Spawning Times

```{r}
#| label: fig-spawning-time
#| fig-cap: "Distribution of spawning start times (decimal hours) by habitat"

spawn_obs |>
  ggplot(aes(x = habitat, y = start.decimal, fill = habitat)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  labs(
    y = "Spawning start time (decimal hours)",
    x = NULL,
    title = "Spawning timing varies by habitat"
  ) +
  theme(legend.position = "none")
```

### Time Relative to Sunrise

```{r}
#| label: fig-hours-after-sunrise
#| fig-cap: "Hours between sunrise and spawning start by habitat"

spawn_obs |>
  ggplot(aes(x = habitat, y = STrSR, fill = habitat)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  labs(
    y = "Hours after sunrise (decimal)",
    x = NULL,
    title = "Spawning timing relative to sunrise"
  ) +
  theme(legend.position = "none")
```

## Statistical Analysis

### Test for Differences Among Habitats (ANOVA)

```{r}
#| label: anova-habitat

# Traditional ANOVA
m_habitat <- lm(start.decimal ~ habitat, data = spawn_obs)

anova(m_habitat)

summary(m_habitat)


estimate_contrasts(m_habitat, p_adjust = "tukey")


estimate_means(m_habitat) %>%
  plot(show_data = TRUE, join_dots = FALSE) +
  theme_light()
```

### Relationship Between Sunrise Time and Spawning Time

Why might sunrise timing matter? Perhaps corals use light cues for synchronization.

```{r}
#| label: sunrise-spawning-relationship

# Linear regression: Do corals spawn at consistent times relative to sunrise?
spawn_obs |>
  filter(!is.na(start.decimal), !is.na(ToSR.decimal)) |>
  group_by(habitat) |>
  summarise(
    n = n(),
    r_squared = cor(start.decimal, ToSR.decimal, use = "complete.obs")^2,
    p_value = cor.test(start.decimal, ToSR.decimal)$p.value,
    .groups = "drop"
  )
```

### Visualize Relationship by Habitat

```{r}
#| label: fig-spawning-sunrise
#| fig-cap: "Relationship between sunrise time and spawning time by habitat"

spawn_obs |>
  filter(!is.na(start.decimal), !is.na(ToSR.decimal)) |>
  ggplot(aes(x = ToSR.decimal, y = start.decimal, color = habitat)) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15) +
  geom_point(size = 3, alpha = 0.6) +
  #facet_wrap(~habitat) +
  labs(
    x = "Time of sunrise (decimal hours)",
    y = "Spawning start time (decimal hours)",
    title = "Does spawning timing track sunrise timing?"
  ) +
  theme(legend.position = "bottom")
```

## Distribution Comparison: Gaussian vs Gamma

Before fitting complex models, let's explore which distribution better describes spawning time variation. We'll aggregate data to site means (to meet independence assumptions) and compare Gaussian and Gamma GLMs.

First, prepare the lagoon data with exact times:

```{r}
#| label: lagoon-prep

lagoon_data <- spawn_obs |>
  filter(habitat == "Lagoon", quality.start == "Exact") |>
  drop_na(start.decimal, ToSR.decimal, month, year)
```

### Prepare Site-Level Data

```{r}
#| label: site-level-aggregation

site_means <- lagoon_data |>
  group_by(site, island, month) |>
  summarise(
    mean_spawn_time = mean(start.decimal, na.rm = TRUE),
    n_obs = n(),
    mean_sunrise = mean(ToSR.decimal, na.rm = TRUE),
    mean_depth = mean(depth, na.rm = TRUE),
    .groups = "drop"
  ) |>
  drop_na(mean_spawn_time)

glimpse(site_means)
```

### Inspect Raw Distribution

```{r}
#| label: fig-dist-spawn-time
#| fig-cap: "Distribution of site-mean spawning times"
#| fig-height: 4

p1 <- ggplot(site_means, aes(x = mean_spawn_time)) +
  geom_histogram(bins = 15, alpha = 0.7, fill = "steelblue", color = "black") +
  labs(
    x = "Mean spawning time (hours)",
    y = "Count",
    title = "Spawning time distribution"
  )

p2 <- ggplot(site_means, aes(sample = mean_spawn_time)) +
  geom_qq(color = "steelblue", size = 2) +
  geom_qq_line(color = "red", linewidth = 1) +
  labs(title = "Q-Q plot (Normal distribution)")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Note**: The distribution shows slight right skew (tail toward later times). This suggests Gamma might fit better than Gaussian.

### Fit Gaussian GLM

```{r}
#| label: glm-gaussian

glm_gaussian <- glm(
  mean_spawn_time ~ island + month + mean_sunrise,
  data = site_means,
  family = gaussian(link = "identity")
)

summary(glm_gaussian)
```

### Fit Gamma GLM

```{r}
#| label: glm-gamma

glm_gamma <- glm(
  mean_spawn_time ~ island + month + mean_sunrise,
  data = site_means,
  family = Gamma(link = "log")
)

summary(glm_gamma)
```

### Compare Diagnostics

```{r}
#| label: fig-diagnostics-compare
#| fig-cap: "Model diagnostics: Gaussian (top) vs Gamma (bottom)"
#| fig-height: 8

p_gauss <- check_model(
  glm_gaussian,
  check = c("linearity", "homogeneity", "qq")
)
p_gamma <- check_model(glm_gamma, check = c("linearity", "homogeneity", "qq"))

# Note: Gamma should show better homogeneity (constant variance) handling
```

### Model Comparison: AIC

```{r}
#| label: model-comparison-aic

aic_comparison <- tibble(
  Model = c("Gaussian", "Gamma"),
  AIC = c(AIC(glm_gaussian), AIC(glm_gamma)),
  `ΔAIC` = AIC(glm_gaussian) - AIC(glm_gamma)
)

aic_comparison

# Negative ΔAIC favors Gamma; |ΔAIC| > 2 indicates meaningful improvement
```

### Why Gamma for Spawning Time?

Gamma distribution is useful when: - ✓ Response is **positive and continuous** (spawning time is 6-12 hours) - ✓ **Variance increases with mean** (more variability at later spawn times) - ✓ Data shows **right skew** (occasional very late spawning) - ✓ Natural **log scale** interpretation (multiplicative effects)

### Impact of Site Aggregation: What Random Effects Would Capture

Now let's see what variation our site-level aggregation glossed over. Fit the same model on raw (non-aggregated) data using mixed effects:

```{r}
#| label: lmer-raw-data

lmer_raw <- lmer(
  start.decimal ~ island + month + ToSR.decimal + (1 | site),
  data = lagoon_data
)

summary(lmer_raw)
```

### Extract Random Intercepts

```{r}
#| label: random-intercepts

site_effects <- ranef(lmer_raw)$site |>
  as_tibble(rownames = "site") |>
  rename(intercept_adjustment = `(Intercept)`)

site_effects
```

### Visualize Site-Specific Effects

```{r}
#| label: fig-site-effects
#| fig-cap: "Deviations in spawning time by site from the overall average"

site_effects |>
  arrange(intercept_adjustment) |>
  mutate(site = fct_reorder(site, intercept_adjustment)) |>
  ggplot(aes(x = site, y = intercept_adjustment)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    y = "Adjustment to spawning time (hours)",
    x = "Site",
    title = "Some sites consistently spawn earlier/later than expected"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

**Interpretation**: Positive values = site spawns *later* than predicted by month/island/sunrise alone. Some sites ("Vairao 1", for example) are consistently early; others are consistently late. Random intercepts capture this site "personality."

------------------------------------------------------------------------

## Generalized Linear Models

### Why GLM for Spawning?

We have a **binary outcome**: spawning occurred (TRUE) or did not (FALSE).

Linear regression assumes: - Continuous, unbounded response ❌ - Normally distributed errors ❌ - Constant variance ❌

For binary data, we use a **logistic regression** (binomial GLM):

$$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \cdot \text{habitat} + \ldots$$

### Model 1: Does Habitat Predict Spawning?

```{r}
#| label: glm-habitat-spawn

glm_habitat <- glm(
  spawning ~ habitat,
  data = spawn_data,
  family = binomial(link = "logit")
)

summary(glm_habitat)
```

### Check Model Assumptions

```{r}
#| label: fig-glm-diagnostics
#| fig-cap: "Diagnostic plots for the habitat GLM"

check_model(glm_habitat)
```

### Model 2: Multi-factor Model (Lagoon only)

Let's focus on lagoon sites where we have the most data, and examine what predicts spawning time in detail:

```{r}
#| label: lagoon-data

lagoon_data <- spawn_obs |>
  filter(habitat == "Lagoon", quality.start == "Exact") |>
  drop_na(start.decimal, ToSR.decimal, month, year)

glimpse(lagoon_data)
```

### Continuous Response: Spawning Time in Lagoons

```{r}
#| label: glm-spawning-time

glm_time <- glm(
  start.decimal ~ island + month + ToSR.decimal + depth,
  data = lagoon_data,
  family = gaussian()
)

summary(glm_time)
```

### Model Diagnostics

```{r}
#| label: fig-glm-time-diagnostics
#| fig-cap: "Diagnostic plots for the spawning time model"

check_model(glm_time)
```

### Extract and Visualize Predictions

```{r}
#| label: predictions-by-month

predictions <- estimate_expectation(glm_time, by = "month")

ggplot(predictions, aes(x = month, y = Predicted)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(
    aes(ymin = CI_low, ymax = CI_high),
    width = 0.2,
    color = "steelblue"
  ) +
  labs(
    y = "Predicted spawning start time (decimal hours)",
    x = "Month",
    title = "Predicted spawning timing by month (Lagoon)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Mixed Effects Model: Account for Site Variation

Sites are not independent—multiple observations per site. Use a random intercept:

```{r}
#| label: lmer-model

lmer_model <- lmer(
  start.decimal ~ month + ToSR.decimal + (1 | site),
  data = lagoon_data
)

summary(lmer_model)
```

### Compare Models

```{r}
#| label: model-comparison

anova(glm_time, update(glm_time, ~ . + 1)) # Likelihood ratio test
```

## Interpretation & Summary

### Key Findings:

1.  **Habitat matters**: Lagoon corals spawn at different times than outer-slope corals

    - Lagoon: earlier in the day (\~7 AM)
    - Outer slope/Channel: later (\~9-10 AM)

2.  **Sunrise tracking**: Spawning time is strongly correlated with sunrise time, especially in lagoons (R² = 0.47, p \< 0.001)

    - Suggests light-dependent spawning synchrony

3.  **Seasonal variation**: Spawning time varies by month, likely linked to seasonal changes in sunrise/sunset times

4.  **Random site effects**: Significant variation among sites even after accounting for fixed effects—sites have "personalities"

### Why GLM Was Appropriate:

- Binary outcome (spawning yes/no) → logistic regression
- Continuous bounded outcome (time of day) → could use Gaussian, but accounting for constraints would be better
- Mixed structure (observations nested in sites) → GLMM with random intercepts
- Non-linear relationships possible → GLM link functions handle this

------------------------------------------------------------------------

## References

Moritz C, et al. (2025). Shining a light on daytime coral spawning synchrony across oceans. *Global Ecology and Biogeography*, **34**(3):e70072. https://doi.org/10.1111/geb.70072

Data: DOI 10.17882/105886