---
title: "Untitled"
format: html
---

## Context

This analysis explores coral (*Porites rus*) spawning timing patterns across different habitats and islands in French Polynesia. We'll investigate whether spawning behavior varies by environmental factors using generalized linear models.

**Reference**: Moritz et al. (2025). Shining a light on daytime coral spawning synchrony across oceans. *Global Ecology and Biogeography*, 10.1111/geb.70072


```{r}
library(here)
library(janitor)
library(tidyverse)
```

```{r}
dat_colony <- read.csv(
  here(
    "lectures/07/Porites_rus_spawning/data/SEANOE_ColonySpawning_Moritz.csv"
  ),
  sep = ";",
  dec = "."
) %>%
  clean_names() %>%
  filter(
    sex %in% c("MALE", "FEMALE"),
    site == "Vairao",
    spawning == "YES"
  ) %>%
  select(-date) %>%
  mutate(
    date = as_date(paste(year, month, day, sep = "-")),
    h_after_sunrise = time - sunrise,
  ) %>%
  filter(!is.na(h_after_sunrise)) %>%
  select(date, site, colony = number, sex, spawning, h_after_sunrise, sst) %>%
  filter(!is.na(h_after_sunrise))

```

```{r}
write_csv(
  dat_colony,
  here(
    "lectures/07/Porites_rus_spawning/data/prus_colony.csv"
  )
)
```


```{r}
dat_sites <- read.csv(
  here(
    "lectures/07/Porites_rus_spawning/data/SEANOE_SiteSpawning_Moritz.csv"
  ),
  sep = ";",
  dec = "."
) %>%
  clean_names() %>%
  mutate(date = as_date(date_us)) %>%
  filter(quality_start == "Exact") %>%
  select(
    -date_us,
    -year,
    -month,
    -quality_start,
    -no_start,
    -s_tr_sr,
    -sst
  ) %>%
  rename(h_start = start_decimal, h_sunrise = to_sr_decimal) %>%
  mutate(h_after_sunrise = h_start - h_sunrise) %>%
  filter(!is.na(h_after_sunrise))
```

```{r}
write_csv(
  dat_sites,
  here(
    "lectures/07/Porites_rus_spawning/data/prus_sites.csv"
  )
)
```
