---
title: Beyond Normality
subtitle: Lecture 5 
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)
theme_set(theme_light(base_size = 20))
```

## Previous lecture recap {.smaller}

::: med-space
:::

-   Linear models with categorical (ANOVA) and numeric (regression) variables
-   Post-hoc tests to compare groups
-   Assumptions of linear models: independence, homoscedasticity, normality
-   Checking assumptions with `check_model()`


## Reminder: Normal distribution {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: false

x <- seq(-4, 4, length.out = 100)
y <- dnorm(x, mean = 0, sd = 1)
df <- data.frame(x, y)

ggplot(df, aes(x, y)) +
  geom_area(fill = "lightblue", alpha = 0.5) +
  geom_line(color = "darkred", linewidth = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
  annotate(
    "segment",
    x = 0,
    xend = 1,
    y = 0.24,
    yend = 0.24,
    arrow = arrow(ends = "both", length = unit(0.2, "cm"))
  ) +
  annotate("text", x = 0.5, y = 0.27, label = "SD", size = 6) +
  annotate("text", x = 0, y = -0.02, label = "Mean", size = 6) +
  labs(x = "Value", y = "Relative frequency") +
  theme(panel.grid = element_blank())
```
:::

::: column
**Properties:**

-   Described by mean $\mu$ and standard deviation $\sigma$
-   Symmetric bell shape
-   Continuously defined from $-\infty$ to $+\infty$
-   Most values close to the mean
:::
:::::

## "Problematic" properties of the normal {.smaller}

::: med-space
:::

::: incremental
1.  **Symmetric**: Equal tails on both sides of the mean
    -   ❌ Problem for asymmetric data

2.  **Defined on the entire real line**: From $-\infty$ to $+\infty$
    -   ❌ Problem if your data cannot be negative
    -   ❌ Problem if your data has an upper limit
    -   ❌ Problem if you have count data

3.  **Constant variance**: Same spread everywhere
    -   ❌ Problem if variability changes with the mean
:::

## Our model so far {.smaller}

::: med-space
:::

For regression (for example):

$$
\begin{align}
y_i &\sim \mathrm{Normal}(\mu_i, \sigma) \\
\mu_i &= a + b \cdot x_i
\end{align}
$$

::: med-space
:::

::: fragment
::: question
But what exactly follows a normal distribution?
:::
:::

## Normality of **residuals** {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: false

set.seed(42)
x_demo <- seq(2, 8, length.out = 20)
y_demo <- 1.5 + 0.8 * x_demo + rnorm(20, 0, 0.8)
df_demo <- data.frame(x = x_demo, y = y_demo)
m_demo <- lm(y ~ x, data = df_demo)

ggplot(df_demo, aes(x = x, y = y)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
  geom_segment(
    aes(x = x, xend = x, y = y, yend = fitted(m_demo)),
    color = "grey40",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  labs(x = "x", y = "y") +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
-   The **residuals** (difference between data and predictions) follow the normal

-   **Not the raw data itself**

-   $\mathrm{residual}_i = y_i - \mu_i$
:::
:::
:::::

::: fragment
::: med-space
:::

The normal distribution describes the **random noise** around the regression line
:::

## When normality fails {.smaller}

::: med-space
:::

Some data types can **never** produce normally distributed residuals:

::: incremental
1.  **Count data**: 0, 1, 2, 3, ... (no negative values, no decimals)
2.  **Binary data**: Yes/No, Presence/Absence, Alive/Dead
3.  **Strictly positive data**: Concentrations, biomass, response times
:::

## Example 1: Sea urchin counts {.smaller}

::: med-space
:::

::::: columns
::: column
You count the number of sea urchins in 1 m² quadrats

```{r}
#| echo: false

set.seed(123)
site <- rep(c("Site A", "Site B"), each = 30)
count_data <- tibble(
  site = site,
  count = c(rpois(30, lambda = 3), rpois(30, lambda = 8))
)

ggplot(count_data, aes(x = count)) +
  geom_histogram(
    binwidth = 1,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  facet_wrap(~site) +
  labs(x = "Number of urchins", y = "Frequency") +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
**Problems with a normal model:**

-   Counts are integers (0, 1, 2, ...)
-   No negative values possible
-   Variance often increases with the mean
-   Asymmetric distribution
:::
:::
:::::

## Example 2: Line intercept transects {.smaller}

::: med-space
:::

::::: columns
::: column
At each of 5 sites, coral cover was estimated with a line intercept transect: along a 20 m transect, at 25 cm intervals (80 points per transect), record if coral is present or absent. Total: 3 transects × 80 points = 240 observations per site

```{r}
#| echo: false

set.seed(456)
site <- rep(1:5, each = 240)
point <- rep(1:80, 5 * 3)
coral_cover <- rep(c(0.15, 0.35, 0.50, 0.25, 0.60), each = 240)
prob_coral <- coral_cover
coral_presence <- rbinom(5 * 240, 1, prob_coral)

df_coral <- tibble(
  site = site,
  point = point,
  presence = coral_presence
)

# Aggregate by site to show the variation
df_coral_summary <- df_coral %>%
  group_by(site) %>%
  summarise(
    coral_cover = mean(presence),
    n_points = n()
  )

ggplot(df_coral_summary, aes(x = as.factor(site), y = coral_cover)) +
  geom_point(size = 3, alpha = 0.6, color = "steelblue") +
  geom_jitter(
    data = df_coral,
    aes(x = as.factor(site), y = presence),
    width = 0.15,
    height = 0.02,
    alpha = 0.1,
    size = 1,
    color = "steelblue"
  ) +
  labs(x = "Site", y = "Coral cover (proportion)") +
  ylim(0, 1) +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
**Problems with a normal model:**

-   Response is a **proportion** (0 to 1)
-   Model could predict values < 0 or > 1
-   Residuals cannot be normal
-   Variance depends on the proportion (higher near 0.5)
:::
:::
:::::

## Example 3: Phytoplankton biomass {.smaller}

::: med-space
:::

::::: columns
::: column
You measure biomass (mg/L) as a function of temperature

```{r}
#| echo: false

set.seed(789)
temp_vals <- seq(15, 30, length.out = 40)
biomass <- exp(0.5 + 0.08 * temp_vals + rnorm(40, 0, 0.3))

df_biomass <- tibble(
  temperature = temp_vals,
  biomass = biomass
)

ggplot(df_biomass, aes(x = temperature, y = biomass)) +
  geom_point(size = 2, alpha = 0.6, color = "steelblue") +
  labs(x = "Temperature (°C)", y = "Biomass (mg/L)") +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
**Problems with a normal model:**

-   Biomass cannot be negative
-   Variance increases with the mean
-   Asymmetric distribution (long tail to the right)
:::
:::
:::::

## The solution: GLMs {.smaller}

::: med-space
:::

::: question
**Generalized Linear Models (GLMs)** extend linear models to handle other distributions
:::

::: med-space
:::

::: incremental
-   **Different distributions**: Poisson, binomial, gamma, etc.
-   **Same concepts**: explanatory variables, coefficients, predictions
-   **New component**: link function
:::

## The link problem {.smaller}

::: med-space
:::

For counts (Poisson distribution), we need $\mu > 0$

::: med-space
:::

::::: columns
::: column
**Classical linear model:**

$$
\mu_i = a + b \cdot x_i
$$

::: fragment
❌ Can produce negative values!
:::
:::

::: column
::: fragment
**GLM with log link:**

$$
\log(\mu_i) = a + b \cdot x_i
$$

✅ $\mu_i = e^{a + b \cdot x_i}$ is always positive!
:::
:::
:::::

## Link functions {.smaller}

::: med-space
:::

The link function transforms the distribution parameter to match the linear model

::: med-space
:::

| Data type | Distribution | Typical link | Ensures |
|-----------------|--------------|--------------|--------|
| Counts | Poisson | log | $\mu > 0$ |
| Binary (0/1) | Binomial | logit | $0 < p < 1$ |
| Strictly positive | Gamma | log | $\mu > 0$ |
| Normal | Normal | identity | no constraint |

## Data generation model with GLM {.smaller}

::: med-space
:::

**Before (linear model):**

$$
\begin{align}
y_i &\sim \mathrm{Normal}(\mu_i, \sigma) \\
\mu_i &= a + b \cdot x_i
\end{align}
$$

::: fragment
**Now (GLM for counts):**

$$
\begin{align}
y_i &\sim \mathrm{Poisson}(\lambda_i) \\
\log(\lambda_i) &= a + b \cdot x_i
\end{align}
$$

or equivalently: $\lambda_i = e^{a + b \cdot x_i}$
:::

## Example: Sea urchin counts with GLM {.smaller}

::: med-space
:::

```{r}
#| echo: true
#| eval: false

# Classical linear model (incorrect)
m_normal <- lm(count ~ site, data = count_data)

# GLM with Poisson distribution (correct)
m_poisson <- glm(
  count ~ site,
  data = count_data,
  family = poisson(link = "log")
)

summary(m_poisson)
```

::: fragment
```{r}
#| echo: false

m_poisson <- glm(
  count ~ site,
  data = count_data,
  family = poisson(link = "log")
)
summary(m_poisson)
```

:::

## Interpreting GLM coefficients {.smaller}

::: med-space
:::

With a log link, coefficients are on the **log scale**:

$$
\log(\lambda_i) = a + b \cdot x_i
$$

::: fragment
To get the effect on the original scale:

$$
\lambda_i = e^{a + b \cdot x_i} = e^a \cdot e^{b \cdot x_i}
$$

A coefficient of $b = 0.5$ means that $\lambda$ is multiplied by $e^{0.5} \approx 1.65$ for each unit increase in $x$
:::

## Checking GLM models {.smaller}

::: med-space
:::

The same tools work!

```{r}
#| echo: true
#| eval: false

library(performance)
check_model(m_poisson)
```


::: fragment
**Key differences:**

-   No normality check (we don't assume normal residuals)
-   Overdispersion check (variance > mean for Poisson)
-   Deviance residuals rather than simple residuals
:::

## Marine example: Coral cover {.smaller}

::: med-space
:::

Compare coral cover across different sites

```{r}
#| echo: true
#| eval: false

# GLM with binomial distribution
m_coral <- glm(
  presence ~ site,
  data = df_coral,
  family = binomial(link = "logit")
)

summary(m_coral)
```

::: fragment
The **logit** link ensures predicted probabilities stay between 0 and 1:

$$
\log\left(\frac{p_i}{1-p_i}\right) = a + b \cdot \text{site}_i
$$
:::

## Visualizing GLM predictions {.smaller}

::: med-space
:::

```{r}
#| echo: false
#| fig-height: 6

m_coral <- glm(
  presence ~ site,
  data = df_coral,
  family = binomial(link = "logit")
)

pred_means <- df_coral %>%
  group_by(site) %>%
  summarise(
    site_num = first(site),
    observed_cover = mean(presence)
  )

pred_means$predicted_prob <- predict(m_coral, newdata = pred_means, type = "response")

ggplot(pred_means, aes(x = as.factor(site_num))) +
  geom_point(
    aes(y = observed_cover),
    size = 3,
    color = "steelblue",
    alpha = 0.7
  ) +
  geom_point(
    aes(y = predicted_prob),
    size = 3,
    color = "darkred",
    alpha = 0.7,
    shape = 1,
    stroke = 2
  ) +
  labs(
    x = "Site",
    y = "Coral cover (probability)",
    title = "Binomial GLM: Observed vs predicted"
  ) +
  ylim(0, 1) +
  theme(panel.grid = element_blank())
```

## When to use which model? {.smaller}

::: med-space
:::

| Response type | Distribution | Marine example |
|-----------------|--------------|---------------|
| Continuous, can be negative | Normal | Temperature, size difference |
| Counts | Poisson | Number of larvae, fish abundance |
| Binary (Yes/No) | Binomial | Survival, presence/absence |
| Strictly positive | Gamma | Biomass, nutrient concentration |

## Summary {.smaller}

::: med-space
:::

::: incremental
1.  The **normal distribution describes residuals**, not raw data
2.  Some data (counts, binary, strictly positive) cannot have normal residuals
3.  **GLMs** extend linear models to other distributions
4.  **Link functions** connect the linear model to the distribution parameter
5.  You already know the concepts — GLMs are just a generalization
:::

## Next week {.smaller}

::: med-space
:::

-   Fitting GLMs in R with `glm()`
-   Interpreting outputs and coefficients
-   Comparing models (Poisson vs binomial vs gamma)
-   Practical examples with marine data

## Your turn {.smaller}

::: med-space
:::

1.  Go to the course website
2.  Go to the calendar
3.  Click on `TP 05`
4.  Connect to Posit Cloud

::: huge
::: center-align

[andieich.github.io/2026_biostatistiques_2](https://andieich.github.io/2026_biostatistiques_2/)

:::
:::
