---
title: Beyond Normality
subtitle: Lecture 5 
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)
library(scales)
library(ggthemes)
library(modelbased)
theme_set(theme_light(base_size = 20))

r2_binomial <- function(model) {
  if (!inherits(model, "glm")) {
    stop("`model` must be a glm object.")
  }
  fam <- stats::family(model)$family
  if (!identical(fam, "binomial")) {
    stop("`model` must use the binomial family.")
  }
  if (is.null(model$null.deviance) || is.null(model$deviance)) {
    stop("Model does not contain deviance values.")
  }

  r2 <- 1 - (model$deviance / model$null.deviance)

  data.frame(
    R2 = r2,
    Method = "Deviance explained",
    stringsAsFactors = FALSE
  )
}

```

## Previous lecture recap {.smaller}

::: med-space
:::

-   Linear models with categorical (ANOVA), numeric (regression), or both (ANCOVA) variables
-   Post-hoc tests to compare groups
-   Assumptions of linear models: independence, homoscedasticity, normality
-   Checking assumptions with `check_model()`


## Reminder: Normal distribution {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: false
#| fig-height: 7

x <- seq(-4, 4, length.out = 100)
y <- dnorm(x, mean = 0, sd = 1)
df <- data.frame(x, y)

ggplot(df, aes(x, y)) +
  geom_area(fill = "lightblue", alpha = 0.5) +
  geom_line(color = "darkred", linewidth = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
  annotate(
    "segment",
    x = 0,
    xend = 1,
    y = 0.24,
    yend = 0.24,
    arrow = arrow(ends = "both", length = unit(0.2, "cm"))
  ) +
  annotate("text", x = 0.5, y = 0.27, label = "SD", size = 6) +
  annotate("text", x = 0, y = -0.02, label = "Mean", size = 6) +
  labs(x = "Value", y = "Relative frequency") +
  theme(panel.grid = element_blank())
```

:::

::: column
**Properties:**

-   Described by mean $\mu$ and standard deviation $\sigma$
-   Symmetric bell shape
-   Continuously defined from $-\infty$ to $+\infty$
-   Most values close to the mean
:::
:::::

## "Problematic" properties of the normal {.smaller}

::: med-space
:::

::: incremental
1.  **Symmetric**: Equal tails on both sides of the mean
    -   ❌ Problem for asymmetric data

2.  **Defined on the entire interval**: From $-\infty$ to $+\infty$
    -   ❌ Problem if your data cannot be negative (weight, length)
    -   ❌ Problem if your data has an upper limit (rates, percent)
    -   ❌ Problem if you have count data

3.  **Constant variance**: Same spread everywhere
    -   ❌ Problem if variability changes with the mean (higher variability for larger individuals)
:::

## Our model so far {.smaller}

::: med-space
:::

For regression (for example):

$$
\begin{align}
y_i &\sim \mathrm{Normal}(\mu_i, \sigma) \\
\mu_i &= a + b \cdot x_i
\end{align}
$$

::: med-space
:::

::: fragment
::: question
But what exactly follows a normal distribution?
:::
:::

## Normality of **residuals** {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: false

set.seed(42)
x_demo <- seq(2, 8, length.out = 20)
y_demo <- 1.5 + 0.8 * x_demo + rnorm(20, 0, 0.8)
df_demo <- data.frame(x = x_demo, y = y_demo)
m_demo <- lm(y ~ x, data = df_demo)

ggplot(df_demo, aes(x = x, y = y)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
  geom_segment(
    aes(x = x, xend = x, y = y, yend = fitted(m_demo)),
    color = "grey40",
    linetype = "11",
    linewidth = 0.8
  ) +
  labs(x = "x", y = "y") +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
-   The **residuals** (difference between data and predictions) follow the normal

-   Not the raw data itself

-   $\mathrm{residual}_i = y_i - \mu_i$
:::
:::
:::::

::: fragment
::: med-space
:::

The normal distribution describes the **random noise** around the regression line
:::

::: fragment
$$
\begin{align}
y_i &\sim \mathrm{Normal}(\mu_i, \sigma) \\
\mu_i &= a + b \cdot x_i
\end{align}
$$

:::

## When normality fails {.smaller}

::: med-space
:::

Some data types can **never** produce normally distributed residuals:

::: incremental
1.  **Count data**: 0, 1, 2, 3, ... (no negative values, no decimals)
2.  **Binary data**: Yes/No, Presence/Absence, Alive/Dead
3.  **Strictly positive data**: Concentrations, biomass, response times
:::

## Example 1: Sea urchin counts {.smaller}

::: med-space
:::

::::: columns
::: column
You count the number of sea urchins in 1 m² quadrats

```{r}
#| echo: false

set.seed(123)
site <- rep(c("Site A", "Site B"), each = 30)
count_data <- tibble(
  site = site,
  count = c(rpois(30, lambda = 3), rpois(30, lambda = 8))
)

ggplot(count_data, aes(x = count)) +
  geom_histogram(
    binwidth = 1,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  facet_wrap(~site) +
  labs(x = "Number of urchins", y = "Frequency") +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
**Problems with a normal model:**

-   Counts are integers (0, 1, 2, ...)
-   No negative values possible
-   Variance often increases with the mean
-   Asymmetric distribution
:::
:::
:::::

## Example 2: Line intercept transects {.smaller}

::: med-space
:::

::::: columns
::: column
**Coral cover**

-  3x 20 m transects
-  25 cm intervals (80 points per transect)
-  Record if coral is present or absent

```{r}
#| echo: false

set.seed(456)
df_coral <- tidyr::expand_grid(
  site = c("A", "B", "C"),
  transect = 1:3,
  point = 1:80
) %>%
  mutate(
    coral_cover = dplyr::case_when(
      site == "A" ~ 0.15,
      site == "B" ~ 0.35,
      site == "C" ~ 0.60
    ),
    presence = rbinom(n(), 1, coral_cover)
  )

# Aggregate by site to show the variation
df_coral_summary <- df_coral %>%
  group_by(site) %>%
  summarise(
    coral_cover = mean(presence),
    n_points = n()
  )

ggplot(df_coral_summary, aes(x = as.factor(site), y = coral_cover)) +
  geom_point(size = 3, alpha = 0.6, color = "steelblue") +
  geom_point(
    data = df_coral,
    aes(x = as.factor(site), y = presence, color = as.factor(transect)),
    position = position_jitterdodge(
      dodge.width = 0.2,
      jitter.width = 0.1,
      jitter.height = 0.02,
    ),
    alpha = 0.1,
    size = 1
  ) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Site", y = "Coral cover") +
  scale_colour_colourblind() +
  theme(panel.grid = element_blank(), legend.position = "none")
```

:::

::: column
::: fragment
**Problems with a normal model:**

-   Response is a **proportion** (0 to 1)
-   Model could predict values < 0 or > 1
-   Residuals cannot be normal
-   Variance depends on the proportion (higher near 0.5)
:::
:::
:::::

## Example 3: Phytoplankton biomass {.smaller}

::: med-space
:::

::::: columns
::: column
You measure biomass (mg/L) as a function of temperature

```{r}
#| echo: false

set.seed(789)
temp_vals <- seq(15, 30, length.out = 40)
biomass <- exp(0.5 + 0.08 * temp_vals + rnorm(40, 0, 0.3))

df_biomass <- tibble(
  temperature = temp_vals,
  biomass = biomass
)

ggplot(df_biomass, aes(x = temperature, y = biomass)) +
  geom_point(size = 2, alpha = 0.6, color = "steelblue") +
  labs(x = "Temperature (°C)", y = "Biomass (mg/L)") +
  theme(panel.grid = element_blank())
```

:::

::: column
::: fragment
**Problems with a normal model:**

-   Biomass cannot be negative
-   Variance increases with the mean
-   Asymmetric distribution (long tail to the right)
:::
:::
:::::

## Consequences {.smaller}

::: med-space
:::


1.  **Impossible predictions**: The model might predict negative counts, or proportions > 100%

::: incremental

2.  **Wrong uncertainty**: Confidence intervals and p-values become unreliable
    -   You might conclude something is significant when it's not
    -   Or miss true patterns in your data

3.  **Biased estimates**: Your slope and intercept will be systematically wrong

4.  **Violated assumptions**: Residual plots show clear patterns instead of random scatter
:::



## The solution: GLMs {.smaller}

::: med-space
:::

::: question
**Generalized Linear Models (GLMs)** extend linear models to handle other distributions
:::

::: med-space
:::

::: incremental
-   **Different distributions**: Poisson, binomial, gamma, etc.
-   **Same concepts**: explanatory variables, coefficients, predictions
:::

::: med-space
:::

::: fragment 
::: {.columns}
::: {.column}
$$
y_i \sim \mathrm{Normal}(\mu, \sigma)
$$

:::
::: {.column}
$$
y_i \sim \mathrm{Gamma}(\alpha, \beta)
$$

::: 
:::
:::





## When to use which distribution? {.smaller}

::: med-space
:::

| Response type | Distribution | Example |
|-----------------|--------------|---------------|
| Continuous | Normal | Temperature, size difference |
| Counts | Poisson | Number of larvae, fish abundance |
| Binary (Yes/No) | Binomial | Survival, presence/absence |
| Strictly positive | Gamma | Biomass, nutrient concentration |



## Distributions and parameters {.smaller}

::: med-space
:::

| Distribution | Formula | Parameters | Constraints |
|---|---|---|---|
| Normal | $y_i \sim \mathrm{Normal}(\mu, \sigma)$ | $\mu$ (mean)<br/>$\sigma$ (SD) | $\mu \in \mathbb{R}$<br/>$\sigma > 0$ |
| Poisson | $y_i \sim \mathrm{Poisson}(\lambda)$ | $\lambda$ (mean) | $\lambda > 0$ |
| Binomial | $y_i \sim \mathrm{Binomial}(n, p)$ | $n$ (trials)<br/>$p$ (prob.) | $n \in \mathbb{N}$<br/>$p \in (0, 1)$ |
| Gamma | $y_i \sim \mathrm{Gamma}(\alpha, \beta)$ | $\alpha$ (shape)<br/>$\beta$ (rate) | $\alpha > 0$<br/>$\beta > 0$ |


## Link functions  {.smaller}

::: med-space
:::

::: {.columns}
::: {.column}
#### Normal Model

One continuous variable ($y$) increases/decreases with other continuous variable ($x$)

:::
::: {.column}
#### Poisson Model
Counts ($y$) increase/decrease with other continuous variable ($x$)

:::
:::
::: {.columns}
::: {.column}

$$
\begin{align}
\mathrm{y}_i \sim \mathrm{Normal(\mu}_i,~\mathrm{\sigma)} \\
\mu_i = a + b \cdot \mathrm{x}_i\\
\end{align}
$$

:::
::: {.column}

$$
\begin{align}
\mathrm{y}_i \sim \mathrm{Poisson}(\lambda_i) \\
\log(\lambda_i) = a + b \cdot \mathrm{x}_i\\
\end{align}
$$

:::
:::

::: {.columns}
::: {.column}
$\mu_i$ can have any value
:::
::: {.column}
$\lambda_i$ has to be > 0!
:::
:::

## Link functions {.smaller}

::: med-space
:::

::: {.columns}
::: {.column}
#### Normal Model

One continuous variable ($y$) increases/decreases with other continuous variable ($x$)

:::
::: {.column}
#### Poisson Model
Counts ($y$) increase/decrease with other continuous variable ($x$)

:::
:::
::: {.columns}
::: {.column}

$$
\begin{align}
\mathrm{y}_i \sim \mathrm{Normal(\mu}_i,~\mathrm{\sigma)} \\
\mu_i = a + b \cdot \mathrm{x}_i\\
\end{align}
$$

:::
::: {.column}

$$
\begin{align}
\mathrm{y}_i \sim \mathrm{Poisson}(\lambda_i) \\
\begin{aligned}
{\color{darkorange}\log}(\lambda_i) &= a + b \cdot \mathrm{x}_i\\
\lambda_i &= e^{a + b \cdot x_i}
\end{aligned}
\end{align}
$$

:::
:::

::: {.columns}
::: {.column}
$\mu_i$ can have any value
:::
::: {.column}
$\lambda_i$ has to be > 0!
:::
:::

::: {.columns}
::: {.column}
:::
::: {.column}


[Link function]{style="color: darkorange;"} ensures that $\lambda_i$  is positive


:::
:::

## Link functions {.smaller}

::: med-space
:::

The link function transforms the distribution parameter to match the linear model

::: med-space
:::

| Data type | Distribution | Typical link | Ensures |
|-----------------|--------------|--------------|--------|
| Counts | Poisson | log | $\mu > 0$ |
| Binary (0/1) | Binomial | logit | $0 < p < 1$ |
| Strictly positive | Gamma | log | $\mu > 0$ |
| Normal | Normal | identity | no constraint |


## Poisson GLM: Sea urchin counts {.smaller}
#### Fit model

::: med-space
:::

```{r}
#| echo: true

# Classical linear model (incorrect)
m_normal <- lm(count ~ site, data = count_data)

# GLM with Poisson distribution (correct)
m_poisson <- glm(
  count ~ site,
  data = count_data,
  family = poisson(link = "log")
)
```


## Poisson GLM: Sea urchin counts {.smaller}
#### Check assumptions

::: med-space
:::

The same tools work!

```{r}
#| echo: true
#| output-location: slide
#| fig-height: 6

library(performance)
check_model(m_poisson)
```


::: fragment
**Key differences:**

-   No normality check (we don't assume normal residuals)
-   Overdispersion check (variance > mean for Poisson)
-   Deviance residuals rather than simple residuals
:::


## Poisson GLM: Sea urchin counts {.smaller}
#### Analysis

::: {.columns}
::: {.column}

Impact explanatory variables:

::: small-output
```{r}
#| echo: true

anova(m_poisson)
```

:::
:::
::: {.column}
::: fragment
Predicted means

::: small-output
```{r}
#| echo: true

estimate_means(m_poisson)
```

:::
:::
:::
:::

::: {.columns}
::: {.column}
::: fragment
Pairwise comparisons (post-hoc test)

::: small-output

```{r}
#| echo: true

estimate_contrasts(m_poisson, p_adjust = "tukey")
```

:::
:::
:::
::: {.column}
::: fragment
$R^2$ value (different to `lm()`!)

::: small-output
```{r}
#| echo: true

r2(m_poisson)
```

::: 
:::
:::
:::

## Poisson GLM: Sea urchin counts {.smaller}
#### Visualisation


::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: true
#| eval: false

#fmt: skip
estimate_means(m_poisson) %>%
  plot()
```

:::
::: {.column width="67%"}

```{r}
estimate_means(m_poisson) %>%
  plot() +
  geom_point(
    position = position_jitter(.2, seed = 444),
    data = count_data,
    aes(x = site, y = count),
    col = "steelblue",
    alpha = 0
  )
```

:::
:::

## Poisson GLM: Sea urchin counts {.smaller}
#### Visualisation


::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: true
#| eval: false

#fmt: skip
estimate_means(m_poisson) %>%
  plot() +
  labs(x = "Site", 
       y = "# Urchins") +
  geom_jitter(
    data = count_data,
    aes(x = site, y = count),
    width = .2,
    col = "steelblue",
    alpha = .3
  )
```


:::
::: {.column width="67%"}

```{r}
estimate_means(m_poisson) %>%
  plot() +
  labs(x = "Site", y = "# Urchins") +
  geom_point(
    position = position_jitter(.2, seed = 444),
    data = count_data,
    aes(x = site, y = count),
    col = "steelblue",
    alpha = 0.3
  )
```

:::
:::

## Binomial GLM: Coral cover {.smaller}
#### Fit model

```{r}
# Aggregate point counts by site
df_coral_pa <- df_coral %>%
  group_by(site, transect) %>%
  summarise(
    coral_points = sum(presence),
    non_coral_points = n() - sum(presence)
  ) %>%
  mutate(site = as.factor(site), n_points = coral_points + non_coral_points)
```

```{r}
#| echo: true

m_coral <- glm(
  cbind(coral_points, non_coral_points) ~ site,
  data = df_coral_pa,
  family = binomial(link = "logit")
)
```


::: fragment
The **logit** link ensures predicted probabilities stay between 0 and 1:

$$
\log\left(\frac{p_i}{1-p_i}\right) = a + b \cdot \text{site}_i
$$

::: callout-note
Using `cbind()` preserves sampling effort: 50% cover from 80 points is more precise than 50% from 8 points!
:::
:::


## Binomial GLM: Coral cover {.smaller}
#### Check assumptions

::: med-space
:::

```{r}
#| echo: true
#| fig-height: 6

check_model(m_coral)
```



## Binomial GLM: Coral cover {.smaller}
#### Analysis

::: {.columns}
::: {.column}

Impact explanatory variables:

::: small-output
```{r}
#| echo: true

anova(m_coral)
```

:::
:::
::: {.column}
::: fragment
Predicted means

::: small-output
```{r}
#| echo: true

estimate_means(m_coral)
```

:::
:::
:::
:::

::: {.columns}
::: {.column}
::: fragment
Pairwise comparisons (post-hoc test)

::: small-output

```{r}
#| echo: true

estimate_contrasts(m_coral, p_adjust = "tukey")
```

:::
:::
:::
::: {.column}
::: fragment
$R^2$ value 

::: small-output
```{r}
#| echo: true

r2_binomial(m_coral)
```

*different to `lm()` and poisson GLM!*

::: 
:::
:::
:::

## Binomial GLM: Coral cover {.smaller}
#### Visualisation


::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: true
#| eval: false

#fmt: skip
estimate_means(m_coral) %>%
  plot()
```

:::
::: {.column width="67%"}

```{r}
estimate_means(m_coral) %>%
  plot() +
  geom_point(
    position = position_jitter(.2, seed = 444),
    data = df_coral_pa,
    aes(x = site, y = coral_points / n_points),
    col = "steelblue",
    alpha = 0
  )
```

:::
:::

## Binomial GLM: Coral cover {.smaller}
#### Visualisation


::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: true
#| eval: false

#fmt: skip
estimate_means(m_coral) %>%
  plot() +
  labs(x = "Site", 
       y = "Coral cover (%)") +
  geom_jitter(
    data = df_coral_pa,
    aes(x = site, 
        y = coral_points / n_points),
    col = "steelblue", 
    alpha = .4
  )
```

:::
::: {.column width="67%"}

```{r}
estimate_means(m_coral) %>%
  plot() +
  labs(x = "Site", y = "Coral cover (%)") +
  geom_point(
    position = position_jitter(.2, seed = 444),
    data = df_coral_pa,
    aes(x = site, y = coral_points / n_points),
    col = "steelblue",
    alpha = .4
  )
```

:::
:::


## Gamma GLM: Phytoplankton biomass {.smaller}
#### Fit model


```{r}
#| echo: true

m_gamma <- glm(
  biomass ~ temperature,
  data = df_biomass,
  family = Gamma(link = "log")
)
```

::: med-space
:::

::: {.fragment}

#### Check model

```{r}
#| echo: true
#| output-location: slide
#| fig-height: 6

check_model(m_gamma)
```
:::

## Gamma GLM: Phytoplankton biomass {.smaller}
#### Analysis

::: {.columns}
::: {.column}

Impact explanatory variables:

::: small-output
```{r}
#| echo: true

anova(m_gamma)
```

:::
:::
::: {.column}
::: fragment
Predicted means

::: small-output
```{r}
#| echo: true
#| eval: false

estimate_relation(m_gamma) 
```

```{r}
#| echo: false
#| eval: true

estimate_relation(m_gamma) %>% 
    head()
```

:::
:::
:::
:::

::: {.columns}
::: {.column}
::: fragment
Pairwise comparisons (post-hoc test)

::: small-output

```{r}
#| echo: true

estimate_slopes(m_gamma, p_adjust = "tukey")
```

:::
:::
:::
::: {.column}
::: fragment
$R^2$ value (different to `lm()`!)

::: small-output
```{r}
#| echo: true

r2(m_gamma)
```

::: 
:::
:::
:::

## Gamma GLM: Phytoplankton biomass {.smaller}
#### Visualisation


::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: true
#| eval: false

#fmt: skip
estimate_relation(m_gamma) %>%
  plot()
```

:::
::: {.column width="67%"}

```{r}
estimate_relation(m_gamma)  %>%
  plot() +
  geom_point(
    data = df_biomass,
    aes(x = temperature, y = biomass),
    col = "steelblue",
    alpha = 0
  )
```

:::
:::

## Gamma GLM: Phytoplankton biomass {.smaller}
#### Visualisation


::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: true
#| eval: false

#fmt: skip
estimate_relation(m_gamma) %>%
  plot() +
  labs(x = "Temperature (°C)", 
       y = "Biomass (mg/L)") +
  geom_point(
    data = df_biomass,
    aes(x = temperature, 
        y = biomass),
    col = "steelblue",
    alpha = .3
  )
```


:::
::: {.column width="67%"}

```{r}
estimate_relation(m_gamma) %>%
  plot() +
  labs(x = "Temperature (°C)", 
       y = "Biomass (mg/L)") +
  geom_point(
    data = df_biomass,
    aes(x = temperature, y = biomass),
    col = "steelblue",
    alpha = 0.3
  )
```

:::
:::

## Summary {.smaller}

::: med-space
:::

::: incremental
1.  The **normal distribution describes residuals**, not raw data
2.  Some data (counts, binary, strictly positive) cannot have normal residuals
3.  **GLMs** extend linear models to other distributions
4.  **Link functions** connect the linear model to the distribution parameter
5.  Examples Poisson, Binomial, Gamma GLM
:::

::: fragment
You already know the concepts — GLMs are just a generalization
:::


## Your turn {.smaller}

::: med-space
:::

1.  Go to the course website
2.  Go to the calendar
3.  Click on `TP 05`
4.  Connect to Posit Cloud

::: huge
::: center-align

[andieich.github.io/2026_biostatistiques_2](https://andieich.github.io/2026_biostatistiques_2/)

:::
:::
