---
title: Post-hoc Tests & Assumptions
subtitle: Lecture 4 
---

```{r}
#| echo: false

library(palmerpenguins)
library(tidyverse)
library(performance)
library(modelbased)
library(emmeans)
library(patchwork)

penguins <- penguins %>%
  na.omit()
```

## Summary from last lecture

-   Linear models as a unified framework
-   Regression: numeric explanatory variable
-   ANOVA: categorical explanatory variable
-   Both are linear models!

## The problem with ANOVA {.smaller}

::: med-space
:::

```{r}
penguins %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_jitter(alpha = .75, shape = 21) +
  geom_pointrange(
    stat = "summary",
    fun.data = mean_sdl,
    show.legend = FALSE,
    shape = 21
  ) +
  labs(
    x = NULL,
    y = "Body mass (g)"
  ) +
  scale_fill_viridis_d() +
  theme_classic(base_size = 20) +
  theme(legend.position = "none")
```

## The problem with ANOVA {.smaller}

::: med-space
:::

```{r}
#| echo: true

m_spec <- lm(body_mass_g ~ species, data = penguins)
anova(m_spec)
```

::: med-space
:::

::: question
ANOVA tells us: "**There is a difference** between species"

But **which** species differs from which other?
:::

## Multiple comparisons {.smaller}

::: med-space
:::

For 3 species, we have 3 possible comparisons:

-   Adelie vs. Chinstrap
-   Adelie vs. Gentoo
-   Chinstrap vs. Gentoo

::: med-space
:::

::: fragment
::: question
Why not just do 3 t-tests?
:::
:::

## Alpha error inflation {.smaller}

::: med-space
:::

::::: columns
::: column
Each test has a 5% error risk (α = 0.05)

::: fragment
**Probability of making no error in 1 test:**

0.95
:::

::: fragment
**Probability of making no error in 3 tests:**

$0.95 \times 0.95 \times 0.95 = 0.857$
:::

::: fragment
**Probability of making at least 1 error:**

$1 - 0.857 = 0.143 = 14.3\%$
:::
:::

::: column
::: fragment
```{r}
n_comparisons <- 1:20
error_rate <- 1 - (0.95^n_comparisons)

data.frame(
  n = n_comparisons,
  error = error_rate
) %>%
  ggplot(aes(x = n, y = error)) +
  geom_line(linewidth = 1.5, color = "darkred") +
  geom_point(size = 3, color = "darkred") +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Number of comparisons",
    y = "Probability of at least 1 error"
  ) +
  theme_classic(base_size = 20)
```

:::
:::
:::::

## Family-wise error rate {.smaller}

::: med-space
:::

**Family-Wise Error Rate (FWER)**: The probability of making at least one type I error in a set of tests

::: med-space
:::

| Number of groups | Number of comparisons | FWER (no correction) |
|-------------------|------------------------|------------------------|
| 2                 | 1                      | 5.0%                   |
| 3                 | 3                      | 14.3%                  |
| 4                 | 6                      | 26.5%                  |
| 5                 | 10                     | 40.1%                  |
| 10                | 45                     | 90.0%                  |

::: med-space
:::

::: question
We need to adjust our p-values to control the FWER
:::

## Bonferroni correction {.smaller}

::: med-space
:::

The simplest method: divide α by the number of tests

$$
\alpha_{\text{adjusted}} = \frac{\alpha}{m}
$$

where $m$ is the number of comparisons

::: med-space
:::

::: fragment
**Example with 3 comparisons:**

$$
\alpha_{\text{adjusted}} = \frac{0.05}{3} = 0.0167
$$

We reject $H_0$ only if $p < 0.0167$
:::

## Calculation: Bonferroni correction {.smaller}

::: med-space
:::

```{r}
#| echo: true

# Pairwise comparisons without correction
pairwise.t.test(
  penguins$body_mass_g,
  penguins$species,
  p.adjust.method = "none"
)
```

## Calculation: Bonferroni correction {.smaller}

::: med-space
:::

```{r}
#| echo: true

# Pairwise comparisons with Bonferroni
pairwise.t.test(
  penguins$body_mass_g,
  penguins$species,
  p.adjust.method = "bonferroni"
)
```

## The problem with Bonferroni {.smaller}

::: med-space
:::

Bonferroni is **too conservative** → loss of statistical power

::: med-space
:::

::::: columns
::: column
**Advantages**

-   Simple to understand
-   Controls FWER well
-   Always works

:::

::: column
**Disadvantages**

-   Too strict with many comparisons
-   Increases type II errors (false negatives)
-   Ignores data structure
:::
:::::

## Other adjustment methods {.smaller}

::: med-space
:::

| Method              | Description                                | When to use           |
|----------------------|--------------------------------------------|----------------------------|
| **Bonferroni**       | Divides α by number of tests              | Simple, few tests       |
| **Holm-Bonferroni**  | Sequential Bonferroni (less strict)      | Alternative to Bonferroni   |
| **Tukey HSD**        | Accounts for all comparisons   | Multiple comparisons     |
| **Benjamini-Hochberg** | Controls false discovery rate | Many tests          |

::: med-space
:::

::: question
But p-values only tell part of the story...
:::

## Why prefer confidence intervals? {.smaller}

::: med-space
:::

::::: columns
::: column
**P-values**

-   Binary (significant or not)
-   Don't give effect magnitude
-   Depend on sample size
-   Don't give uncertainty

:::

::: column
**Confidence intervals**

-   Show effect magnitude
-   Show uncertainty
-   Allow biological interpretation
-   Facilitate comparison
:::
:::::

::: med-space
:::

::: fragment
::: question
CIs tell us: "The true difference is probably in this range"
:::
:::

## Pairwise differences with CIs {.smaller}

::: med-space
:::

```{r}
#| echo: true

# Calculate estimated means and comparisons
emm <- emmeans(m_spec, ~species)
pairs_result <- pairs(emm)
pairs_result
```

## Visualizing pairwise differences {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: true
#| eval: false

pairs(emm) %>%
  confint() %>%
  as.data.frame() %>%
  ggplot(aes(x = contrast, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_pointrange(
    aes(ymin = lower.CL, ymax = upper.CL)
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mass difference (g)"
  ) +
  theme_classic(base_size = 18)
```

:::

::: column
```{r}
pairs(emm) %>%
  confint() %>%
  as.data.frame() %>%
  ggplot(aes(x = contrast, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(
    aes(ymin = lower.CL, ymax = upper.CL),
    size = 1,
    linewidth = 1.5
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mass difference (g)"
  ) +
  theme_classic(base_size = 18)
```

:::
:::::

## Using `modelbased` for contrasts {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: true
#| eval: false

library(modelbased)

# Estimate means
estimate_means(m_spec)
```

::: med-space
:::

```{r}
#| echo: true
#| eval: false

# Estimate contrasts
estimate_contrasts(m_spec)
```

:::

::: column
::: small-output
```{r}
estimate_means(m_spec)
```

:::

::: med-space
:::

::: small-output
```{r}
estimate_contrasts(m_spec)
```

:::
:::
:::::

## Visualizing with `modelbased` {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: true
#| eval: false

# Means with CIs
plot(estimate_means(m_spec))
```

::: med-space
:::

```{r}
#| echo: true
#| eval: false

# Contrasts with CIs
plot(estimate_contrasts(m_spec))
```

:::

::: column
```{r}
estimate_means(m_spec) %>%
  as.data.frame() %>%
  ggplot(aes(x = species, y = Mean)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high)) +
  theme_classic(base_size = 16) +
  labs(y = "Body mass (g)", x = "Species")
```

::: med-space
:::

```{r}
estimate_contrasts(m_spec) %>%
  as.data.frame() %>%
  ggplot(aes(x = Level1, y = Difference)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high)) +
  theme_classic(base_size = 16) +
  labs(y = "Mass difference (g)", x = "Contrast")
```

:::
:::::

## CI overlap {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
estimate_means(m_spec) %>%
  as.data.frame() %>%
  ggplot(aes(x = species, y = Mean)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high)) +
  theme_classic(base_size = 20) +
  labs(y = "Body mass (g)", x = NULL)
```

:::

::: column
::: fragment
::: question
**Rule of thumb:**

If the CIs of two means don't overlap, they are probably different (p < 0.05)
:::
:::

::: fragment
::: callout-warning
## Warning!

CI overlap doesn't always imply no significant difference. Contrasts are more precise.
:::
:::
:::
:::::

## Slope comparisons: ANCOVA {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
penguins %>%
  ggplot(aes(
    x = bill_length_mm,
    y = body_mass_g,
    color = species,
    fill = species
  )) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", alpha = 0.2) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(
    x = "Bill length (mm)",
    y = "Body mass (g)",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "bottom")
```

:::

::: column
::: question
Are the slopes different between species?
:::

::: fragment
Model with interaction:

```{r}
#| echo: true

m_ancova <- lm(
  body_mass_g ~ bill_length_mm * species,
  data = penguins
)
```

:::
:::
:::::

## Testing slope differences {.smaller}

::: med-space
:::

::: small-output
```{r}
#| echo: true

summary(m_ancova)
```

:::

::: fragment
::: question
The interaction terms test whether slopes differ from the reference slope (Adelie)
:::
:::

## Visualizing slope differences {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: true
#| eval: false

# Estimate slopes
estimate_slopes(m_ancova, trend = "bill_length_mm", at = "species")
```

:::

::: column
::: small-output
```{r}
estimate_slopes(m_ancova, trend = "bill_length_mm", at = "species")
```

:::
:::
:::::

::: med-space
:::

```{r}
# Visualize slopes by species
penguins %>%
  ggplot(aes(
    x = bill_length_mm,
    y = body_mass_g,
    color = species,
    fill = species
  )) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", alpha = 0.2) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(
    x = "Bill length (mm)",
    y = "Body mass (g)",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "bottom")
```

## Slope comparisons {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| echo: true
#| eval: false

# Compare slopes
estimate_contrasts(
  m_ancova,
  contrast = "species",
  at = "bill_length_mm"
)
```

:::

::: column
::: small-output
```{r}
estimate_contrasts(
  m_ancova,
  contrast = "species",
  at = "bill_length_mm"
)
```

:::
:::
:::::

::: fragment
::: question
Gentoo has a significantly higher slope than the other two species
:::
:::

## Summary: Which method to use? {.smaller}

::: med-space
:::

```{r}
#| echo: false

library(gt)

tibble(
  Situation = c(
    "Planned comparisons",
    "All pairwise comparisons",
    "Few comparisons (< 5)",
    "Many comparisons (> 10)",
    "Interest in effect magnitude",
    "Scientific publication"
  ),
  Method = c(
    "t-tests with Holm correction",
    "Tukey HSD",
    "Bonferroni or Holm",
    "Benjamini-Hochberg",
    "Confidence intervals",
    "CIs + adjusted p-values"
  )
) %>%
  gt() %>%
  tab_options(
    table.font.size = 18,
    data_row.padding = px(8)
  )
```

## Part 2: Linear model assumptions {background-color="#440154"}

::: med-space
:::

::: center-align
::: huge
**When can we use a linear model?**
:::
:::

## The three key assumptions {.smaller}

::: med-space
:::

A linear model assumes that:

::: med-space
:::

::::: columns
::: column
### 1. Independence

Observations are independent of each other

:::

::: column
### 2. Homoscedasticity

Residual variance is constant

:::
:::::

::: med-space
:::

### 3. Normality

Residuals follow a normal distribution

::: med-space
:::

::: fragment
::: question
But first: Why do we need linear models?
:::
:::

## Nonparametric tests vs. Linear models {.smaller}

::: med-space
:::

::::: columns
::: column
**Nonparametric tests**

-   Mann-Whitney, Kruskal-Wallis, etc.
-   "No assumptions"
-   Use ranks

::: fragment
**Problems:**

-   Difficult to interpret biologically
-   No effect magnitude
-   No complex models
-   Less power
-   No estimates/predictions
:::
:::

::: column
::: fragment
**Linear models**

-   Estimate biological parameters
-   Magnitude and uncertainty
-   Extensible (interactions, covariates)
-   More power
-   Allow predictions

::: fragment
::: question
**Robust** to moderate assumption violations!
:::
:::
:::
:::
:::::

## Robustness of linear models {.smaller}

::: med-space
:::

Linear models are surprisingly **robust**:

::: med-space
:::

-   **Normality** is not critical with n > 30 (central limit theorem)
-   **Homoscedasticity**: moderate violation often acceptable
-   **Independence**: the only truly critical one

::: med-space
:::

::: fragment
::: question
**Key message:** Don't abandon linear models too quickly!

They are often better than nonparametric alternatives, even with moderate violations.
:::
:::

## Checking assumptions with `performance` {.smaller}

::: med-space
:::

```{r}
#| echo: true

library(performance)

# Our ANOVA model
m_spec <- lm(body_mass_g ~ species, data = penguins)

# Check all assumptions
check_model(m_spec)
```

## Interpreting diagnostics {.smaller}

::: med-space
:::

::::: columns
::: column
### Posterior Predictive Check

Compares observed data to model predictions

::: fragment
### Linearity

Residuals vs. fitted values (should be flat)
:::

::: fragment
### Homogeneity of Variance

Constant variance of residuals
:::
:::

::: column
::: fragment
### Influential Observations

Points with high influence (Cook's distance)
:::

::: fragment
### Normality of Residuals

**The Q-Q plot** (quantile-quantile)
:::

::: fragment
### Collinearity

Correlation between predictors (not relevant here)
:::
:::
:::::

## The Q-Q plot explained {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| fig-height: 8

check_normality(m_spec) %>%
  plot()
```

:::

::: column
**Quantile-Quantile plot**

-   Compares observed quantiles to theoretical quantiles of a normal distribution
-   If points follow the line: normal distribution
-   Deviations at extremes: heavy/light tails
-   S-pattern: skewness

::: fragment
::: question
More useful than a histogram for assessing normality!
:::
:::
:::
:::::

## Q-Q plot: Typical patterns {.smaller}

::: med-space
:::

```{r}
set.seed(123)
n <- 200

# Normal data
normal_data <- rnorm(n)

# Heavy-tailed data
heavy_tail <- rt(n, df = 3)

# Skewed data
skewed <- rexp(n, rate = 1)

# Create Q-Q plots
p1 <- data.frame(sample = normal_data) %>%
  ggplot(aes(sample = sample)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal distribution", x = "Theoretical", y = "Observed") +
  theme_classic(base_size = 14)

p2 <- data.frame(sample = heavy_tail) %>%
  ggplot(aes(sample = sample)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Heavy tails", x = "Theoretical", y = "Observed") +
  theme_classic(base_size = 14)

p3 <- data.frame(sample = skewed) %>%
  ggplot(aes(sample = sample)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Positive skew", x = "Theoretical", y = "Observed") +
  theme_classic(base_size = 14)

p1 + p2 + p3
```

## Formal normality tests {.smaller}

::: med-space
:::

```{r}
#| echo: true

# Shapiro-Wilk test
check_normality(m_spec)
```

::: med-space
:::

::: fragment
::: callout-warning
## Problem with normality tests

-   **Too sensitive** with large samples (detect trivial deviations)
-   **Not sensitive enough** with small samples (when we'd need them)
-   Don't tell **if** the deviation is biologically important

:::
:::

::: fragment
::: question
**Prefer visual examination (Q-Q plot)** over formal tests!
:::
:::

## "Eyeballing" vs. Statistical tests {.smaller}

::: med-space
:::

::::: columns
::: column
**Formal tests (Shapiro-Wilk)**

-   Answer "yes" or "no"
-   Sensitivity depends on n
-   No information on severity
-   Trivial deviations become "significant"

:::

::: column
**Visual inspection (Q-Q plot)**

-   Shows the type of deviation
-   Shows where it occurs
-   Allows assessing severity
-   Develops statistical intuition
-   **Context**: model robustness

:::
:::::

::: med-space
:::

::: fragment
::: question
**Practical advice:** Use the Q-Q plot as your primary tool. Formal tests are secondary.
:::
:::

## Assumption 1: Independence {.smaller}

::: med-space
:::

**Observations are independent**

::: med-space
:::

Violation = one observation influences another

::: med-space
:::

::::: columns
::: column
**Examples of violations**

-   Repeated measures on the same individuals
-   Temporal autocorrelation
-   Spatial autocorrelation
-   Pseudoreplication
-   Hierarchical structure (sites, populations)

:::

::: column
::: fragment
**Consequences**

-   Underestimation of standard error
-   Confidence intervals too narrow
-   P-values too small
-   **This is the most serious violation!**

:::
:::
:::::

::: fragment
::: question
Solution: Mixed models, time series, spatial models
:::
:::

## Independence violations in ecology {.smaller}

::: med-space
:::

::::: columns
::: column
### Spatial autocorrelation

```{r}
set.seed(42)
n_sites <- 50
x_coord <- runif(n_sites, 0, 100)
y_coord <- runif(n_sites, 0, 100)

# Create simple spatial autocorrelation
spatial_effect <- numeric(n_sites)
for (i in 1:n_sites) {
  distances <- sqrt((x_coord[i] - x_coord)^2 + (y_coord[i] - y_coord)^2)
  weights <- exp(-distances / 20)
  spatial_effect[i] <- sum(rnorm(n_sites, 0, 1) * weights) / sum(weights)
}

response <- 10 + spatial_effect + rnorm(n_sites, 0, 0.5)

spatial_data <- data.frame(
  x = x_coord,
  y = y_coord,
  value = response
)

ggplot(spatial_data, aes(x = x, y = y, color = value)) +
  geom_point(size = 4) +
  scale_color_viridis_c() +
  labs(
    title = "Spatial autocorrelation",
    x = "X coordinate",
    y = "Y coordinate",
    color = "Value"
  ) +
  theme_classic(base_size = 16)
```

Nearby sites are more similar
:::

::: column
::: fragment
### Temporal autocorrelation

```{r}
set.seed(42)
n_time <- 100
time <- 1:n_time

# Time series with autocorrelation
ar_process <- arima.sim(n = n_time, list(ar = 0.7))

temporal_data <- data.frame(
  time = time,
  value = as.numeric(ar_process)
)

ggplot(temporal_data, aes(x = time, y = value)) +
  geom_line(linewidth = 1, color = "darkblue") +
  geom_point(size = 2, color = "darkblue") +
  labs(
    title = "Temporal autocorrelation",
    x = "Time",
    y = "Value"
  ) +
  theme_classic(base_size = 16)
```

Consecutive measurements are correlated
:::
:::
:::::

## Assumption 2: Homoscedasticity {.smaller}

::: med-space
:::

**Constant variance of residuals**

::: med-space
:::

::::: columns
::: column
```{r}
set.seed(123)
x <- runif(100, 0, 10)
y_homo <- 2 + 3 * x + rnorm(100, 0, 2)

data.frame(x = x, y = y_homo) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Homoscedasticity",
    x = "Predictor",
    y = "Response"
  ) +
  theme_classic(base_size = 16)
```

Constant variance ✓
:::

::: column
```{r}
y_hetero <- 2 + 3 * x + rnorm(100, 0, 0.3 * x)

data.frame(x = x, y = y_hetero) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Heteroscedasticity",
    x = "Predictor",
    y = "Response"
  ) +
  theme_classic(base_size = 16)
```

Variance increases ✗
:::
:::::

## Heteroscedasticity in ecology {.smaller}

::: med-space
:::

**Very common in ecological data!**

::: med-space
:::

::::: columns
::: column
**Examples**

-   Variance increases with the mean
-   Count data (abundance)
-   Allometric processes
-   Body size and metabolism

:::

::: column
::: fragment
**Consequences**

-   Biased standard errors
-   CIs and tests less reliable
-   **Less serious** than non-independence

::: question
Solutions: Transformation, GLM, robust models
:::
:::
:::
:::::

## Assumption 3: Normality of residuals {.smaller}

::: med-space
:::

**Residuals (not raw data!) follow a normal distribution**

::: med-space
:::

::::: columns
::: column
::: callout-important
## Warning!

It's **not** the distribution of raw data that must be normal, but that of the **residuals**!
:::

::: fragment
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of raw data",
    x = "Body mass (g)",
    y = "Frequency"
  ) +
  theme_classic(base_size = 16)
```

:::
:::

::: column
::: fragment
```{r}
residuals_spec <- residuals(m_spec)

data.frame(residuals = residuals_spec) %>%
  ggplot(aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of residuals",
    x = "Residuals",
    y = "Frequency"
  ) +
  theme_classic(base_size = 16)
```

:::
:::
:::::

## Non-normal data in ecology {.smaller}

::: med-space
:::

**Data types that often violate normality:**

::: med-space
:::

| Data type    | Problem              | Appropriate distribution |
|--------------------|-----------------------|-------------------------|
| Counts          | Integers, skewed | Poisson, Negative binomial |
| Presence/Absence   | Binary (0/1)         | Binomial               |
| Proportions        | Limited to [0, 1]     | Beta, Binomial         |
| Survival times    | Censored             | Weibull, Exponential  |
| Positive data  | Skewed          | Gamma, Log-normal      |

::: med-space
:::

::: fragment
::: question
**It's normal for your ecological data to be non-normal!**

Classic linear models aren't always appropriate → GLM
:::
:::

## Common violations in ecology {.smaller}

::: med-space
:::

::::: columns
::: column
### Count data

```{r}
set.seed(123)
count_data <- rpois(200, lambda = 3)

data.frame(counts = count_data) %>%
  ggplot(aes(x = counts)) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(
    title = "Poisson distribution",
    x = "Number of individuals",
    y = "Frequency"
  ) +
  theme_classic(base_size = 14)
```

Integers, skewed, variance = mean
:::

::: column
::: fragment
### Proportion data

```{r}
set.seed(123)
prop_data <- rbeta(200, 2, 5)

data.frame(proportions = prop_data) %>%
  ggplot(aes(x = proportions)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
    title = "Beta distribution",
    x = "Proportion",
    y = "Frequency"
  ) +
  theme_classic(base_size = 14)
```

Limited between 0 and 1
:::
:::
:::::

## Important message {.smaller}

::: med-space
:::

::: huge
::: center-align
**At least one assumption is generally violated in ecological data**
:::
:::

::: med-space
:::

::: fragment
::: question
This is **normal** and **expected**!
:::
:::

::: med-space
:::

::: fragment
**What this means:**

1.  Don't panic when assumptions aren't perfect
2.  Assess the **severity** of the violation
3.  Linear models are often **robust** to moderate violations
4.  For serious violations → **Generalized Linear Models (GLM)**

:::

## Introduction to GLMs {.smaller}

::: med-space
:::

**Generalized Linear Models (GLM)**: Extension of linear models

::: med-space
:::

::::: columns
::: column
**Classic linear models**

$$
\begin{align}
Y_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 X_i
\end{align}
$$

-   Normal distribution only
-   Constant variance
-   Direct linear relationship

:::

::: column
::: fragment
**GLM**

$$
\begin{align}
Y_i &\sim \text{Distribution}(\mu_i, \theta) \\
g(\mu_i) &= \beta_0 + \beta_1 X_i
\end{align}
$$

-   Different distributions (Poisson, binomial, gamma, etc.)
-   Variance can depend on the mean
-   Link function $g()$

:::
:::
:::::

::: fragment
::: question
**Next week:** How and when to use GLMs!
:::
:::

## Summary {.smaller}

::: med-space
:::

**Post-hoc tests**

-   Alpha error inflation is a real problem
-   P-value correction (Bonferroni, Tukey, etc.)
-   **CIs are often more informative** than p-values
-   `modelbased` facilitates estimation and visualization

::: med-space
:::

::: fragment
**Linear model assumptions**

-   Independence (critical), homoscedasticity, normality
-   Linear models are **robust** to moderate violations
-   **Better than nonparametric tests** in most cases
-   Use `performance::check_model()` for diagnostics
-   **Q-Q plots > formal tests** to assess normality
-   Violations are **normal** in ecology → GLM

:::

## Your turn {.smaller}

::: med-space
:::

1.  Go to the course website
2.  Go to the calendar
3.  Click on `TP 04`
4.  Connect to Posit Cloud

::: huge
::: center-align
[andieich.github.io/2026_biostatistiques_2](https://andieich.github.io/2026_biostatistiques_2/)
:::
:::
