---
title: Post-hoc Tests & Assumptions
subtitle: Lecture 4 
---

```{r}
#| echo: false

library(palmerpenguins)
library(tidyverse)
library(emmeans)
library(performance)
library(ggthemes)
library(modelbased)
library(patchwork)
options(modelbased_backend = "emmeans")

penguins <- penguins %>%
  na.omit()
```

## Summary from last lecture

-   Linear models as a unified framework:
    -   *Regression*: numeric explanatory variable
    -   *ANOVA/t-test*: categorical explanatory variables
    -   *ANCOVA*: numeric and categorical explanatory variables

## The problem with ANOVA {.smaller}

::: med-space
:::

```{r}
penguins %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_jitter(alpha = .75, shape = 21) +
  geom_pointrange(
    stat = "summary",
    fun.data = mean_sdl,
    show.legend = FALSE,
    shape = 21
  ) +
  labs(
    x = NULL,
    y = "Body mass (g)"
  ) +
  scale_fill_colorblind() +
  theme_classic(base_size = 20) +
  theme(legend.position = "none")
```

## The problem with ANOVA {.smaller}

::: med-space
:::

```{r}
#| echo: true

m_spec <- lm(body_mass_g ~ species, data = penguins)
anova(m_spec)
```

::: med-space
:::

::: question
ANOVA tells us: "**There is a difference** between species"

But **which** species differs from which other?
:::

## Multiple comparisons {.smaller}

::: med-space
:::

For 3 species, we have 3 possible comparisons:

-   Adelie vs. Chinstrap
-   Adelie vs. Gentoo
-   Chinstrap vs. Gentoo

::: med-space
:::

::: fragment
::: question
Why not just do 3 models?
:::
:::

## Alpha error inflation {.smaller}

::: med-space
:::

::::: columns
::: column
-  Each test has a 5% error risk (α = 0.05)

::: fragment
-  **Probability of making no error in 1 test:** 0.95
:::

::: fragment
-  **Probability of making no error in 3 tests:** $0.95 \times 0.95 \times 0.95 = 0.857$
:::

::: fragment
-   **Probability of making at least 1 error:** $1 - 0.857 = 0.143 = 14.3\%$
:::
:::

::: column
::: fragment

::: med-space
:::

```{r}
#| fig-height: 10

n_comparisons <- 1:20
error_rate <- 1 - (0.95^n_comparisons)

data.frame(
  n = n_comparisons,
  error = error_rate
) %>%
  ggplot(aes(x = n, y = error)) +
  geom_line(linewidth = 1.5, color = "darkred") +
  geom_point(size = 3, color = "darkred") +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Number of comparisons",
    y = "Probability of at least 1 error"
  ) +
  theme_classic(base_size = 23)
```

:::
:::
:::::

## Family-wise error rate (FWER) {.smaller}

::: med-space
:::

> 
**FWER:** The probability of making at least one type I error in a set of tests

::: med-space
:::

::: small
| Number of groups | Number of comparisons | FWER (no correction) |
|-------------------|------------------------|------------------------|
| 2                 | 1                      | 5.0%                   |
| 3                 | 3                      | 14.3%                  |
| 4                 | 6                      | 26.5%                  |
| 5                 | 10                     | 40.1%                  |
| 10                | 45                     | 90.0%                  |
:::

::: med-space
:::

::: question
We need to adjust our p-values to control the FWER
:::

## p-value adjustments {.smaller}

::: med-space
:::
::: small
| Method              | Description                                | When to use           |
|----------------------|--------------------------------------------|----------------------------|
| **Bonferroni**       | Divides α by number of tests              | Simple, few tests       |
| **Holm-Bonferroni**  | Sequential Bonferroni (less strict)      | Alternative to Bonferroni   |
| **Benjamini-Hochberg** | Controls false discovery rate | Many tests          |
| **Tukey HSD**        | Accounts for all comparisons   | **Multiple comparisons**     |
:::

::: med-space
:::

::: question
Tukey tests are often used for post-hoc tests 
:::

## Post-hoc Tests in `R` {.smaller}

::: med-space
:::


```{r}
#| echo: true

m_spec <- lm(body_mass_g ~ species, data = penguins)

estimate_contrasts(m_spec, p_adjust = "tukey")
```


## Visualise Post-Hoc Tests {.smaller}

::: med-space
:::


::: {.columns}
::: {.column}
```{r}
#| echo: true
#| eval: false

estimate_means(m_spec, by = "species") %>%
  plot()
```

```{r}
#| echo: false
#| eval: true

estimate_means(m_spec, by = "species") %>%
  plot() +
  theme_light(base_size = 21)
```


:::
::: {.column .small}


#### Recommendation

-   Plot 95% CI

    >
   "**95% Confidence Interval:** is a range calculated such that, if repeated many times, 95% of such intervals would contain the true parameter value."

-   **No overlap:** Groups are statistiaclly significantly different (p > .05)

-   **Overlap:** May be statistically significantly different, but effect is likely small
:::
:::

::: med-space
:::

::: {.fragment .question .small}
Recently, researchers are moving away from p-values as hard cutoffs. Effect size (biological importance) matters more than p < 0.05!
:::


## Slope comparisons: ANCOVA {.smaller}

::: med-space
:::

::::: columns
::: column

::: med-space
:::
```{r}
#| fig-height: 10

penguins %>%
  ggplot(aes(
    x = bill_length_mm,
    y = body_mass_g,
    color = species,
    fill = species
  )) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", alpha = 0.2) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(
    x = "Bill length (mm)",
    y = "Body mass (g)",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "bottom")
```

:::

::: column
::: med-space
:::
#### Different slopes between species?

::: fragment
::: medium-output
```{r}
#| echo: true

m_ancova <- lm(
  body_mass_g ~ bill_length_mm * species,
  data = penguins
)

anova(m_ancova)
```

::: 
:::
:::
:::::


## Slope comparisons: ANCOVA {.smaller}

::: med-space
:::

::::: columns
::: column

::: med-space
:::
```{r}
#| fig-height: 10

penguins %>%
  ggplot(aes(
    x = bill_length_mm,
    y = body_mass_g,
    color = species,
    fill = species
  )) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", alpha = 0.2) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(
    x = "Bill length (mm)",
    y = "Body mass (g)",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "bottom")
```

:::

::: column
::: med-space
:::
#### Different slopes between species?

::: tiny-output
```{r}
#| echo: true

estimate_contrasts(
  m_ancova,
  contrast = "species",
  trend = "bill_length_mm"
)
```

:::

:::
:::::

## Slope comparisons: ANCOVA {.smaller}

::: med-space
:::

::::: columns
::: column

::: med-space
:::
```{r}
#| fig-height: 10

penguins %>%
  ggplot(aes(
    x = bill_length_mm,
    y = body_mass_g,
    color = species,
    fill = species
  )) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", alpha = 0.2) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(
    x = "Bill length (mm)",
    y = "Body mass (g)",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "bottom")
```

:::

::: column
::: med-space
:::
#### Slopes different to 0?

::: medium-output
```{r}
#| echo: true

estimate_slopes(m_ancova, by = "species")


```

:::
:::
:::::


## Test

```{r}

lm(body_mass_g ~ species * sex, data = penguins) %>%
  estimate_means(contrast = c("species", "sex")) %>%
  plot()

lm(body_mass_g ~ species * sex, data = penguins) %>%
  estimate_contrasts(contrast = c("species", "sex"))

lm(body_mass_g ~ species * sex, data = penguins) %>%
  estimate_contrasts(by = c("sex", "species"))


estimate_contrasts(
  lm(body_mass_g ~ species * sex, data = penguins),
  contrast = c("species", "sex")
)
```

## Part 2: Linear model assumptions {background-color="#440154"}

::: med-space
:::

::: center-align
::: huge
**When can we use a linear model?**
:::
:::

## The three key assumptions {.smaller}

::: med-space
:::

A linear model assumes that:

::: med-space
:::

::::: columns
::: column
### 1. Independence

Observations are independent of each other

:::

::: column
### 2. Homoscedasticity

Residual variance is constant

:::
:::::

::: med-space
:::

### 3. Normality

Residuals follow a normal distribution

::: med-space
:::

::: fragment
::: question
But first: Why do we need linear models?
:::
:::

## Nonparametric tests vs. Linear models {.smaller}

::: med-space
:::

::::: columns
::: column
**Nonparametric tests**

-   Mann-Whitney, Kruskal-Wallis, etc.
-   "No assumptions"
-   Use ranks

::: fragment
**Problems:**

-   Difficult to interpret biologically
-   No effect magnitude
-   No complex models
-   Less power
-   No estimates/predictions
:::
:::

::: column
::: fragment
**Linear models**

-   Estimate biological parameters
-   Magnitude and uncertainty
-   Extensible (interactions, covariates)
-   More power
-   Allow predictions

::: fragment
::: question
**Robust** to moderate assumption violations!
:::
:::
:::
:::
:::::

## Robustness of linear models {.smaller}

::: med-space
:::

Linear models are surprisingly **robust**:

::: med-space
:::

-   **Normality** is not critical with n > 30 (central limit theorem)
-   **Homoscedasticity**: moderate violation often acceptable
-   **Independence**: the only truly critical one

::: med-space
:::

::: fragment
::: question
**Key message:** Don't abandon linear models too quickly!

They are often better than nonparametric alternatives, even with moderate violations.
:::
:::

## Checking assumptions with `performance` {.smaller}

::: med-space
:::

```{r}
#| echo: true

library(performance)

# Our ANOVA model
m_spec <- lm(body_mass_g ~ species, data = penguins)

# Check all assumptions
check_model(m_spec)
```

## Interpreting diagnostics {.smaller}

::: med-space
:::

::::: columns
::: column
### Posterior Predictive Check

Compares observed data to model predictions

::: fragment
### Linearity

Residuals vs. fitted values (should be flat)
:::

::: fragment
### Homogeneity of Variance

Constant variance of residuals
:::
:::

::: column
::: fragment
### Influential Observations

Points with high influence (Cook's distance)
:::

::: fragment
### Normality of Residuals

**The Q-Q plot** (quantile-quantile)
:::

::: fragment
### Collinearity

Correlation between predictors (not relevant here)
:::
:::
:::::

## The Q-Q plot explained {.smaller}

::: med-space
:::

::::: columns
::: column
```{r}
#| fig-height: 8

check_normality(m_spec) %>%
  plot()
```

:::

::: column
**Quantile-Quantile plot**

-   Compares observed quantiles to theoretical quantiles of a normal distribution
-   If points follow the line: normal distribution
-   Deviations at extremes: heavy/light tails
-   S-pattern: skewness

::: fragment
::: question
More useful than a histogram for assessing normality!
:::
:::
:::
:::::

## Q-Q plot: Typical patterns {.smaller}

::: med-space
:::

```{r}
set.seed(123)
n <- 200

# Normal data
normal_data <- rnorm(n)

# Heavy-tailed data
heavy_tail <- rt(n, df = 3)

# Skewed data
skewed <- rexp(n, rate = 1)

# Create Q-Q plots
p1 <- data.frame(sample = normal_data) %>%
  ggplot(aes(sample = sample)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal distribution", x = "Theoretical", y = "Observed") +
  theme_classic(base_size = 14)

p2 <- data.frame(sample = heavy_tail) %>%
  ggplot(aes(sample = sample)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Heavy tails", x = "Theoretical", y = "Observed") +
  theme_classic(base_size = 14)

p3 <- data.frame(sample = skewed) %>%
  ggplot(aes(sample = sample)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Positive skew", x = "Theoretical", y = "Observed") +
  theme_classic(base_size = 14)

p1 + p2 + p3
```

## Formal normality tests {.smaller}

::: med-space
:::

```{r}
#| echo: true

# Shapiro-Wilk test
check_normality(m_spec)
```

::: med-space
:::

::: fragment
::: callout-warning
## Problem with normality tests

-   **Too sensitive** with large samples (detect trivial deviations)
-   **Not sensitive enough** with small samples (when we'd need them)
-   Don't tell **if** the deviation is biologically important

:::
:::

::: fragment
::: question
**Prefer visual examination (Q-Q plot)** over formal tests!
:::
:::

## "Eyeballing" vs. Statistical tests {.smaller}

::: med-space
:::

::::: columns
::: column
**Formal tests (Shapiro-Wilk)**

-   Answer "yes" or "no"
-   Sensitivity depends on n
-   No information on severity
-   Trivial deviations become "significant"

:::

::: column
**Visual inspection (Q-Q plot)**

-   Shows the type of deviation
-   Shows where it occurs
-   Allows assessing severity
-   Develops statistical intuition
-   **Context**: model robustness

:::
:::::

::: med-space
:::

::: fragment
::: question
**Practical advice:** Use the Q-Q plot as your primary tool. Formal tests are secondary.
:::
:::

## Assumption 1: Independence {.smaller}

::: med-space
:::

**Observations are independent**

::: med-space
:::

Violation = one observation influences another

::: med-space
:::

::::: columns
::: column
**Examples of violations**

-   Repeated measures on the same individuals
-   Temporal autocorrelation
-   Spatial autocorrelation
-   Pseudoreplication
-   Hierarchical structure (sites, populations)

:::

::: column
::: fragment
**Consequences**

-   Underestimation of standard error
-   Confidence intervals too narrow
-   P-values too small
-   **This is the most serious violation!**

:::
:::
:::::

::: fragment
::: question
Solution: Mixed models, time series, spatial models
:::
:::

## Independence violations in ecology {.smaller}

::: med-space
:::

::::: columns
::: column
### Spatial autocorrelation

```{r}
set.seed(42)
n_sites <- 50
x_coord <- runif(n_sites, 0, 100)
y_coord <- runif(n_sites, 0, 100)

# Create simple spatial autocorrelation
spatial_effect <- numeric(n_sites)
for (i in 1:n_sites) {
  distances <- sqrt((x_coord[i] - x_coord)^2 + (y_coord[i] - y_coord)^2)
  weights <- exp(-distances / 20)
  spatial_effect[i] <- sum(rnorm(n_sites, 0, 1) * weights) / sum(weights)
}

response <- 10 + spatial_effect + rnorm(n_sites, 0, 0.5)

spatial_data <- data.frame(
  x = x_coord,
  y = y_coord,
  value = response
)

ggplot(spatial_data, aes(x = x, y = y, color = value)) +
  geom_point(size = 4) +
  scale_color_viridis_c() +
  labs(
    title = "Spatial autocorrelation",
    x = "X coordinate",
    y = "Y coordinate",
    color = "Value"
  ) +
  theme_classic(base_size = 16)
```

Nearby sites are more similar
:::

::: column
::: fragment
### Temporal autocorrelation

```{r}
set.seed(42)
n_time <- 100
time <- 1:n_time

# Time series with autocorrelation
ar_process <- arima.sim(n = n_time, list(ar = 0.7))

temporal_data <- data.frame(
  time = time,
  value = as.numeric(ar_process)
)

ggplot(temporal_data, aes(x = time, y = value)) +
  geom_line(linewidth = 1, color = "darkblue") +
  geom_point(size = 2, color = "darkblue") +
  labs(
    title = "Temporal autocorrelation",
    x = "Time",
    y = "Value"
  ) +
  theme_classic(base_size = 16)
```

Consecutive measurements are correlated
:::
:::
:::::

## Assumption 2: Homoscedasticity {.smaller}

::: med-space
:::

**Constant variance of residuals**

::: med-space
:::

::::: columns
::: column
```{r}
set.seed(123)
x <- runif(100, 0, 10)
y_homo <- 2 + 3 * x + rnorm(100, 0, 2)

data.frame(x = x, y = y_homo) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Homoscedasticity",
    x = "Predictor",
    y = "Response"
  ) +
  theme_classic(base_size = 16)
```

Constant variance ✓
:::

::: column
```{r}
y_hetero <- 2 + 3 * x + rnorm(100, 0, 0.3 * x)

data.frame(x = x, y = y_hetero) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Heteroscedasticity",
    x = "Predictor",
    y = "Response"
  ) +
  theme_classic(base_size = 16)
```

Variance increases ✗
:::
:::::

## Heteroscedasticity in ecology {.smaller}

::: med-space
:::

**Very common in ecological data!**

::: med-space
:::

::::: columns
::: column
**Examples**

-   Variance increases with the mean
-   Count data (abundance)
-   Allometric processes
-   Body size and metabolism

:::

::: column
::: fragment
**Consequences**

-   Biased standard errors
-   CIs and tests less reliable
-   **Less serious** than non-independence

::: question
Solutions: Transformation, GLM, robust models
:::
:::
:::
:::::

## Assumption 3: Normality of residuals {.smaller}

::: med-space
:::

**Residuals (not raw data!) follow a normal distribution**

::: med-space
:::

::::: columns
::: column
::: callout-important
## Warning!

It's **not** the distribution of raw data that must be normal, but that of the **residuals**!
:::

::: fragment
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of raw data",
    x = "Body mass (g)",
    y = "Frequency"
  ) +
  theme_classic(base_size = 16)
```

:::
:::

::: column
::: fragment
```{r}
residuals_spec <- residuals(m_spec)

data.frame(residuals = residuals_spec) %>%
  ggplot(aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of residuals",
    x = "Residuals",
    y = "Frequency"
  ) +
  theme_classic(base_size = 16)
```

:::
:::
:::::

## Non-normal data in ecology {.smaller}

::: med-space
:::

**Data types that often violate normality:**

::: med-space
:::

| Data type    | Problem              | Appropriate distribution |
|--------------------|-----------------------|-------------------------|
| Counts          | Integers, skewed | Poisson, Negative binomial |
| Presence/Absence   | Binary (0/1)         | Binomial               |
| Proportions        | Limited to [0, 1]     | Beta, Binomial         |
| Survival times    | Censored             | Weibull, Exponential  |
| Positive data  | Skewed          | Gamma, Log-normal      |

::: med-space
:::

::: fragment
::: question
**It's normal for your ecological data to be non-normal!**

Classic linear models aren't always appropriate → GLM
:::
:::

## Common violations in ecology {.smaller}

::: med-space
:::

::::: columns
::: column
### Count data

```{r}
set.seed(123)
count_data <- rpois(200, lambda = 3)

data.frame(counts = count_data) %>%
  ggplot(aes(x = counts)) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(
    title = "Poisson distribution",
    x = "Number of individuals",
    y = "Frequency"
  ) +
  theme_classic(base_size = 14)
```

Integers, skewed, variance = mean
:::

::: column
::: fragment
### Proportion data

```{r}
set.seed(123)
prop_data <- rbeta(200, 2, 5)

data.frame(proportions = prop_data) %>%
  ggplot(aes(x = proportions)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
    title = "Beta distribution",
    x = "Proportion",
    y = "Frequency"
  ) +
  theme_classic(base_size = 14)
```

Limited between 0 and 1
:::
:::
:::::

## Important message {.smaller}

::: med-space
:::

::: huge
::: center-align
**At least one assumption is generally violated in ecological data**
:::
:::

::: med-space
:::

::: fragment
::: question
This is **normal** and **expected**!
:::
:::

::: med-space
:::

::: fragment
**What this means:**

1.  Don't panic when assumptions aren't perfect
2.  Assess the **severity** of the violation
3.  Linear models are often **robust** to moderate violations
4.  For serious violations → **Generalized Linear Models (GLM)**

:::

## Introduction to GLMs {.smaller}

::: med-space
:::

**Generalized Linear Models (GLM)**: Extension of linear models

::: med-space
:::

::::: columns
::: column
**Classic linear models**

$$
\begin{align}
Y_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 X_i
\end{align}
$$

-   Normal distribution only
-   Constant variance
-   Direct linear relationship

:::

::: column
::: fragment
**GLM**

$$
\begin{align}
Y_i &\sim \text{Distribution}(\mu_i, \theta) \\
g(\mu_i) &= \beta_0 + \beta_1 X_i
\end{align}
$$

-   Different distributions (Poisson, binomial, gamma, etc.)
-   Variance can depend on the mean
-   Link function $g()$

:::
:::
:::::

::: fragment
::: question
**Next week:** How and when to use GLMs!
:::
:::

## Summary {.smaller}

::: med-space
:::

**Post-hoc tests**

-   Alpha error inflation is a real problem
-   P-value correction (Bonferroni, Tukey, etc.)
-   **CIs are often more informative** than p-values
-   `modelbased` facilitates estimation and visualization

::: med-space
:::

::: fragment
**Linear model assumptions**

-   Independence (critical), homoscedasticity, normality
-   Linear models are **robust** to moderate violations
-   **Better than nonparametric tests** in most cases
-   Use `performance::check_model()` for diagnostics
-   **Q-Q plots > formal tests** to assess normality
-   Violations are **normal** in ecology → GLM

:::

## Your turn {.smaller}

::: med-space
:::

1.  Go to the course website
2.  Go to the calendar
3.  Click on `TP 04`
4.  Connect to Posit Cloud

::: huge
::: center-align
[andieich.github.io/2026_biostatistiques_2](https://andieich.github.io/2026_biostatistiques_2/)
:::
:::


## Add this

### Data is not independend

Actually, the data on coral cover I shared with you looks different. There are three transects per site! 
What is the problem if you test for the coastal orienation?
The data is not anymore independend! The ecological conditions within each site are more similar than between the sites, you have a structrue in your data!

I real life, we can solve this problem by telling R the structure of the data. These are called random effect models. Not explained in more detail here

==> Mixed models, read about "random factors" if you are interested.


### Variance is homogenious

Similary, if variance differs between groups, or increases with one variable, one can tell the model to consider this. 

==> Variabce structure (not covered in this course, read about it if you want)


### Normality
Finally the residuls are often not normally distributed, e.g. bacuase you have stricly positive data (weight, height), integers (counts), or percent data (only between 0 a 100%). A normal distributiio is defined betwwn - Inf an + Inf, and does not fit to these data types. This leads to wron p-values! So it is important to consider. We will cover this in the next lectures and learn when to use what
==> GLMs

